(ref:tidyversepart) Data Science with `tidyverse`

```{r echo=FALSE, results="asis", purl=FALSE}
cat("# (PART) Data Science with tidyverse {-} ")
```
# Data Visualization {#viz}

```{r setup-viz, include=FALSE, purl=FALSE}
# Used to define Learning Check numbers:
chap <- 2
lc <- 0
```


We begin the development of your data science toolbox with data visualization. 
By visualizing data, 
we gain valuable insights we couldn't initially obtain 
from just looking at the raw data values.  
We'll use the `ggplot2` package, 
as it provides an easy way to customize your plots. 
`ggplot2` is rooted in the data visualization theory 
known as _the grammar of graphics_ [@wilkinson2005], 
developed by Leland Wilkinson. \index{Wilkinson, Leland}

At their most basic, graphics/plots/charts 
(we use these terms interchangeably in this book) 
provide a nice way to explore the patterns in data, 
such as the presence of *outliers*, *distributions* of individual variables, 
and *relationships* between groups of variables. 
Graphics are designed to emphasize the findings and insights 
you want your audience to understand. 
This does, however, require a balancing act. 
On the one hand, 
you want to highlight as many interesting findings as possible. 
On the other hand, you don't want to include so much information 
that it overwhelms your audience.  

As we will see, plots \index{plots} also help us to identify patterns 
and outliers in our data. 
We'll see that a common extension of these ideas is 
to compare the *distribution* \index{distribution} of one continuous variable, 
such as what are the center and spread of the values, 
as we go across the levels of a different categorical variable. 


### Needed packages {-}

We will be playing with a few demo datasets throughout this chapter. 
The following code gives us access to the data, 
as well as the to some tools so we can interact with the data. 

```{r load-package, eval=F}
# Install xfun so that I can use xfun::pkg_load2
if (!requireNamespace('xfun')) install.packages('xfun')
xf <- loadNamespace('xfun')

cran_primary <- c(
                  "dplyr", 
                  "gapminder", 
                  "ggplot2", 
                  "nycflights13", 
                  "tibble"
)

if (length(cran_primary) != 0) xf$pkg_load2(cran_primary)

gg <- import::from(ggplot2, .all=TRUE, .into={new.env()})
dp <- import::from(dplyr, .all=TRUE, .into={new.env()})
import::from(magrittr, '%>%')
```

```{r import-pkg, echo=F, message=FALSE, warning=FALSE}
cran_secondary <- c(
                    "kableExtra", 
                    "patchwork", 
                    "readr", 
                    "stringr"
)

gg <- import::from(ggplot2, .all=TRUE, .into={new.env()})
dp <- import::from(dplyr, .all=TRUE, .into={new.env()})
import::from(magrittr, '%>%')
# need `patchwork` for `<gg> + <gg>` to work, 
# cannot explicitly import `+` separately, so importing everything
import::from(patchwork, .all=T)
```


## The grammar of graphics {#grammarofgraphics}

We start with a discussion of a theoretical framework for data visualization 
known as "the grammar of graphics." 
This framework serves as the foundation for the \index{R packages!ggplot2} 
`ggplot2` package which we'll use extensively in this chapter. 
\index{Grammar of Graphics, The} 
Think of how we construct and form sentences in English 
by combining different elements, 
like nouns, verbs, articles, subjects, objects, etc. 
We can't just combine these elements in any arbitrary order; 
we must do so following a set of rules known as a linguistic grammar. 
Similarly to a linguistic grammar, 
"the grammar of graphics" defines a set of rules 
for constructing *statistical graphics* 
by combining different types of *layers*. 
This grammar was created by Leland Wilkinson [@wilkinson2005] 
and has been implemented in a variety of data visualization software platforms 
like R, but also [Plotly](https://plot.ly/) 
and [Tableau](https://www.tableau.com/).

### Components of the grammar

In short, the grammar tells us that:

> **A statistical graphic is a `mapping` of `data` variables 
to `aes`thetic attributes of `geom`etric objects.**

Specifically, we can break a graphic 
into the following three essential components:

1. `data`: the dataset containing the variables of interest.

1. `geom`: the geometric object in question. 
   This refers to the type of object we can observe in a plot. 
   For example: points, lines, and bars.

1. `aes`: aesthetic attributes of the geometric object. 
   For example, x/y position, color, shape, and size.  
   Aesthetic attributes are *mapped* to variables in the dataset.

You might be wondering why we wrote the terms `data`, `geom`, and `aes` 
in a computer code type font. 
We'll see very shortly that we'll specify the elements of the grammar in R 
using these terms. 
However, let's first break down the grammar with an example.


### Gapminder data {#gapminder}

```{r echo=FALSE, purl=FALSE}
gapminder_2007 <- gapminder::gapminder %>%
  dp$filter(year == 2007) %>%
  dp$select(-year) %>%
  dp$rename(
    Country = country,
    Continent = continent,
    `Life Expectancy` = lifeExp,
    `Population` = pop,
    `GDP per Capita` = gdpPercap
  )
```


In February 2006, 
a Swedish physician and data advocate named Hans Rosling 
gave a TED talk titled 
["The best stats you've ever seen"](https://www.ted.com/talks/hans_rosling_shows_the_best_stats_you_ve_ever_seen) 
where he presented global economic, health, and development data 
from the website 
[gapminder.org](http://www.gapminder.org/tools/#_locale_id=en;&chart-type=bubbles). 
For example, for data on `r gapminder_2007 %>% nrow()` countries in 2007, 
let's consider only a few countries in Table \@ref(tab:gapminder-2007) 
as a peek into the data.

```{r gapminder-2007, echo=FALSE, purl=FALSE}
gapminder_2007 %>%
  head(3) %>%
  knitr::kable(
    digits = 2,
    caption = "Gapminder 2007 Data: First 3 of 142 countries"
  ) %>%
  kableExtra::kable_styling(
    font_size = ifelse(knitr::is_latex_output(), 10, 16),
    latex_options = c("hold_position")
  )
```

Each row in this table corresponds to a country in 2007. 
For each row, we have `r gapminder_2007 %>% ncol()` columns:

1. **Country**: Name of country.

1. **Continent**: Which of the five continents the country is part of. 
   Note that "Americas" includes countries in both North and South America 
   and that Antarctica is excluded.

1. **Life Expectancy**: Life expectancy in years.

1. **Population**: Number of people living in the country.

1. **GDP per Capita**: Gross domestic product (in US dollars).

Now consider Figure \@ref(fig:gapminder), 
which plots this for all `r gapminder_2007 %>% nrow()` of the data's countries.

```{r gapminder, echo=FALSE, fig.cap="Life expectancy over GDP per capita in 2007.", purl=FALSE}
gapminder_plot <- gg$ggplot(
  data = gapminder_2007,
  mapping = gg$aes(
    x = `GDP per Capita`,
    y = `Life Expectancy`,
    size = Population,
    color = Continent
  )
) +
  gg$geom_point() +
  gg$labs(x = "GDP per capita", y = "Life expectancy")

gapminder_plot
```

Let's view this plot through the grammar of graphics:

1. The `data` variable **GDP per Capita** gets mapped 
   to the `x`-position `aes`thetic \index{ggplot2!aes()} of the points.

1. The `data` \index{ggplot2!data} variable **Life Expectancy** gets mapped 
   to the `y`-position `aes`thetic of the points.

1. The `data` variable **Population** gets mapped 
   to the `size` `aes`thetic of the points.

1. The `data` variable **Continent** gets mapped 
   to the `color` `aes`thetic of the points.

We'll see shortly that `data` corresponds to the particular data frame 
where our data is saved and 
that "data variables" correspond to particular columns in the data frame. 
Furthermore, the type of `geom`etric object \index{ggplot2!geom} considered 
in this plot are points. 
That being said, while in this example we are considering points, 
graphics are not limited to just points. 
We can also use lines, bars, and other geometric objects.

Let's summarize the three essential components of the grammar 
in Table \@ref(tab:summary-table-gapminder).

```{r summary-table-gapminder, echo=FALSE, purl=FALSE}
tibble::tibble(
  `data variable` = 
    c("GDP per Capita", "Life Expectancy", "Population", "Continent"),
  aes = c("x", "y", "size", "color"),
  geom = c("point", "point", "point", "point")
) %>%
  knitr::kable(
    caption = "Summary of the grammar of graphics for this plot",
    booktabs = TRUE,
    linesep = ""
  ) %>%
  kableExtra::kable_styling(
    font_size = ifelse(knitr::is_latex_output(), 10, 16),
    latex_options = c("hold_position")
  )
```


### Other components

There are other components of the grammar of graphics we can control as well.  
As you start to delve deeper into the grammar of graphics, 
you'll start to encounter these topics more frequently. 
In this book, we'll keep things simple 
and only work with these two additional components:

- `facet`ing breaks up a plot into several plots split 
  by the values of another variable 
  (Section \@ref(facets)) \index{ggplot2!facet}
- `position` adjustments for barplots 
  (Section \@ref(geombar)) \index{ggplot2!position}

Other more complex components like `scales` and `coord`inate systems 
are left for a more advanced text such as 
[*R for Data Science*](http://r4ds.had.co.nz/data-visualisation.html#aesthetic-mappings) 
[@rds2016]. 
Generally speaking, the grammar of graphics 
allows for a high degree of customization of plots 
and also a consistent framework for easily updating and modifying them.


### ggplot2 package

In this book, we will use the `ggplot2` package for data visualization, 
which is an implementation of the `g`rammar of `g`raphics for R 
[@R-ggplot2]. 
As we noted earlier, a lot of the previous section was written 
in a computer code type font. 
This is because the various components of the grammar of graphics 
are specified in the `ggplot()` \index{ggplot2!ggplot()} function 
included in the `ggplot2` package. 
For the purposes of this book, 
we'll always provide the `ggplot()` function with the following arguments 
(i.e., inputs) at a minimum:

* The data frame where the variables exist: the `data` argument.
* The mapping of the variables to aesthetic attributes: 
  the `mapping` argument which specifies the `aes`thetic attributes involved.

After we've specified these components, 
we then add *layers* to the plot using the `+` sign. 
The most essential layer to add to a plot is the layer 
that specifies which type of `geom`etric object we want the plot to involve: 
points, lines, bars, and others. 
Other layers we can add to a plot include the plot title, 
axes labels, visual themes for the plots, 
and facets (which we'll see in Section \@ref(facets)).

Let's now put the theory of the grammar of graphics into practice.

## Five named graphs - the 5NG {#FiveNG}

In order to keep things simple in this book, 
we will only focus on five different types of graphics, 
each with a commonly given name. 
We term these "five named graphs" or in abbreviated form, 
the **5NG**: \index{five named graphs}

1. scatterplots 
1. linegraphs
1. histograms
1. boxplots 
1. barplots

We'll also present some variations of these plots, 
but with this basic repertoire of five graphics in your toolbox, 
you can visualize a wide array of different variable types. 
Note that certain plots are only appropriate for categorical variables, 
while others are only appropriate for continuous variables. 



## 5NG#1: Scatterplots {#scatterplots}

The simplest of the 5NG are *scatterplots*, \index{scatterplots} 
also called *bivariate plots*. 
They allow you to visualize the *relationship* between two continuous variables. 
While you may already be familiar with scatterplots, 
let's view them through the lens of the grammar of graphics 
we presented in Section \@ref(grammarofgraphics). 
Specifically, we will visualize the relationship 
between the following two continuous variables in the `flights` data frame 
included in the \index{R packages!nycflights13} `nycflights13` package:

1. `dep_delay`: departure delay on the horizontal "x" axis and
1. `arr_delay`: arrival delay on the vertical "y" axis



```{r df_flights, message=F, warning=F, echo=FALSE, purl=FALSE}
# This redundant code is used for dynamic non-static in-line text output purposes
import::from(nycflights13, df_flights = flights)
flights_rows <- df_flights %>%
  nrow() %>%
  formatC(big.mark = ",")
#   scales::comma()
alaska_flights_rows <- df_flights %>%
  dp$filter(carrier == "AS") %>%
  nrow() %>%
  formatC(big.mark = ",")
#   comma()
# flights_cols <- df_flights %>% ncol()
```

As we did before in Chapter \@ref(nycflights13), 
let's first import `flights` and save it as `df_flights`. 

```{r eval=F}
import::from(nycflights13, df_flights = flights)
```

Next, let's pare down the data from all `r flights_rows` flights 
that left NYC in 2013, 
to only the `r alaska_flights_rows` *Alaska Airlines* flights 
that left NYC in 2013. 
We do this so our scatterplot will involve 
a manageable `r alaska_flights_rows` points, 
and not an overwhelmingly large number like `r flights_rows`. 

To achieve this, 
we'll take the newly imported `df_flights` data frame, 
filter the rows so that only the `r alaska_flights_rows` rows 
corresponding to Alaska Airlines flights are kept, 
and save this in a new data frame called `df_alaska_flights` 
using the `<-` *assignment* operator: \index{operators!assignment (<-)} 

```{r}
df_alaska_flights <- df_flights %>% 
  dplyr::filter(carrier == "AS")
```

This code above uses the `dplyr` package for data wrangling to achieve our goal: 
it takes the `df_flights` data frame 
and `filter`s it to only return the rows 
where `carrier` is equal to `"AS"`, 
Alaska Airlines' carrier code. 
Recall from Section \@ref(code) that 
testing for equality is specified with \index{using == instead of =} `==` 
and not `=`. 
This `filter`ing action is one type of **data wrangling**, 
a topic we will discuss in detail in Chapter \@ref(wrangling). 

Now try it yourself 
and explore the resulting data frame by running `View(df_alaska_flights)`. 
You'll see that it has `r alaska_flights_rows` rows, 
consisting of only `r alaska_flights_rows` Alaska Airlines flights. 
Now you should be convinced that the code above 
achieved what it was supposed to. 

### Scatterplots via `geom_point` {#geompoint}

Let's now go over the code that will create the desired scatterplot, 
while keeping in mind the grammar of graphics framework 
we introduced in Section \@ref(grammarofgraphics). 
Let's take a look at the code and break it down piece-by-piece.


```{r eval=FALSE}
gg$ggplot(data = df_alaska_flights, 
          mapping = gg$aes(x = dep_delay, y = arr_delay)) + 
  gg$geom_point()
```

Within the `ggplot()` \index{ggplot2!ggplot()} function, 
we specify two of the components of the grammar of graphics 
as arguments (i.e., inputs):

1. The `data` as the `df_alaska_flights` data frame 
   via `data = df_alaska_flights`.
1. The `aes`thetic \index{ggplot2!mapping} `mapping` 
   by setting `mapping = gg$aes(x = dep_delay, y = arr_delay)`. 
   Specifically, the variable `dep_delay` maps to the `x` position aesthetic, 
   while the variable `arr_delay` maps to the `y` position.
        
We then add a layer to the `ggplot()` function call using the `+` sign. 
The added layer in question specifies the third component of the grammar: 
the `geom`etric object. 
In this case, the geometric object is set to be points 
by specifying `geom_point()`. 
After running these two lines of code in your console, 
you'll notice two outputs: 
a warning message and the graphic shown in Figure \@ref(fig:noalpha).

```{r noalpha, fig.cap="Arrival delays versus departure delays for Alaska Airlines flights from NYC in 2013.", warning=TRUE, echo=FALSE, purl=FALSE}
gg$ggplot(data = df_alaska_flights, 
          mapping = gg$aes(x = dep_delay, y = arr_delay)) +
  gg$geom_point()
```

Let's first unpack the graphic in Figure \@ref(fig:noalpha). 
Observe that a *positive relationship* 
exists between `dep_delay` and `arr_delay`: 
as departure delays increase, arrival delays tend to also increase.  
Observe also the large mass of points clustered near (0, 0), 
the point indicating flights that neither departed nor arrived late. 

Let's turn our attention to the warning message. 
R is alerting us to the fact that five rows were ignored 
due to them being missing. 
For these 5 rows, either the value for `dep_delay` or `arr_delay` 
or both were missing (recorded in R as `NA`), 
and thus these rows were ignored in our plot.
Let's confirm that there are indeed five rows 
with missing data by identifying those rows. 

```{r eval = F}
dp$filter(df_alaska_flights, is.na(dep_delay) | is.na(arr_delay))
```

```{r missing-data, echo=F}
dp$filter(df_alaska_flights, is.na(dep_delay) | is.na(arr_delay)) %>% 
  knitr::kable(caption = "Five rows with missing data in `dep_delay` and/or `arr_delay`") %>% 
	kableExtra::kable_styling(bootstrap_options = 
                            c("striped", "hover", "condensed", "responsive"), 
                          font_size = 14) %>% 
  kableExtra::column_spec(c(6,9), bold=T, color="white", background="#D7261E") %>% 
  kableExtra::scroll_box(width = "100%")
```


Sure enough, the five row all have missing data in one or the other columns. 
(Place your mouse cursor inside the table and scroll right 
to see more columns.) 

Before we continue,
let's make a few more observations about the code 
that created the scatterplot. 
Note that the `+` sign comes at the end of lines, and not at the beginning.
You'll get an error in R if you put it at the beginning of a line. 
\index{ggplot2!+} 
When adding layers to a plot, 
you are encouraged to start a new line *after* the `+` 
(by pressing the Return/Enter button on your keyboard) 
so that the code for each layer is on a new line. 
As we add more and more layers to plots, 
you'll see this will greatly improve the legibility of your code.

To stress the importance of adding the layer specifying the `geom`etric object, 
consider Figure \@ref(fig:nolayers) where no layers are added. 
Because the `geom`etric object was not specified, 
we have a blank plot which is not very useful!

```{r nolayers, fig.cap="A plot with no layers."}
gg$ggplot(data = df_alaska_flights, 
          mapping = gg$aes(x = dep_delay, y = arr_delay))
```


```{block lc-scatterplots, type="learncheck", purl=FALSE}
\vspace{-0.15in}
**_Learning check_**
\vspace{-0.1in}
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What are some practical reasons why `dep_delay` and `arr_delay` 
have a positive relationship?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What variables in the `nycflights13::weather` data frame would you expect 
to have a negative correlation (i.e., a negative relationship) 
with `dep_delay`? Why? 
Remember that we are focusing on continuous variables here. 
Hint: Explore the `nycflights13::weather` dataset by using the `View()` function. 

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
Why do you believe there is a cluster of points near (0, 0)? 
What does (0, 0) correspond to in terms of the Alaska Air flights?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What are some other features of the plot that stand out to you?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
Create a new scatterplot using different variables 
in the `df_alaska_flights` data frame by modifying the example given.

```{block, type="learncheck", purl=FALSE}
\vspace{-0.25in}
\vspace{-0.25in}
```


### Overplotting {#overplotting}

<!-- I need a better example to show how jitter could be useful.  -->

The large mass of points near (0, 0) in Figure \@ref(fig:noalpha) 
can cause some confusion 
since it is hard to tell the true number of points that are plotted. 
This is the result of a phenomenon called \index{overplotting} *overplotting*. 
As one may guess, this corresponds to points being plotted 
on top of each other over and over again. 
When overplotting occurs,
it is difficult to know the number of points being plotted. 
There are two methods to address the issue of overplotting. 
Either by 

1. Adjusting the transparency of the points or
1. Adding a little random "jitter", or random "nudges", to each of the points.

**Method 1: Changing the transparency**

The first way of addressing overplotting is to change the transparency/opacity 
of the points by setting the `alpha` argument in `geom_point()`. 
We can change the `alpha` argument to be any value between `0` and `1`,
where `0` sets the points to be 100% transparent 
and `1` sets the points to be 100% opaque. 
By default, `alpha` is set to `1`. 
In other words, if we don't explicitly set an `alpha` value, 
R will use `alpha = 1`.

Note how the following code is identical to the code in 
Section \@ref(scatterplots) that created the scatterplot with overplotting, 
but with `alpha = 0.2` added to the `geom_point()` function:

```{r alpha, fig.cap="Arrival vs. departure delays scatterplot with alpha = 0.2."}
gg$ggplot(data = df_alaska_flights, 
          mapping = gg$aes(x = dep_delay, y = arr_delay)) + 
  gg$geom_point(alpha = 0.2)
```

The key feature to note in Figure \@ref(fig:alpha) is that 
the transparency \index{ggplot2!alpha} \index{adding transparency to plots} 
of the points is cumulative: 
areas with a high-degree of overplotting are darker, 
whereas areas with a lower degree are less dark. 

Note furthermore that there is no `aes()` surrounding `alpha = 0.2`. 
This is because we are not mapping a variable to an aesthetic attribute, 
but rather merely changing the default setting of `alpha`. 
In fact, you'll receive an error if you try to change the second line 
to read `geom_point(aes(alpha = 0.2))`.

**Method 2: Jittering the points**

The second way of addressing overplotting is by *jittering* all the points. 
This means giving each point a small "nudge" in a random direction. 
You can think of "jittering" as shaking the points around a bit on the plot. 
Let's illustrate using a simple example first. 
Say we have a data frame with 4 identical rows of x and y values: 
(0,0), (0,0), (0,0), and (0,0). 
In Figure \@ref(fig:jitter-example-plot-1), 
we present both the regular scatterplot of these 4 points (on the left) 
and its jittered counterpart (on the right). 

```{r jitter-example-plot-1, fig.cap="Regular and jittered scatterplot.", out.width="80%", echo=FALSE, purl=FALSE}
jitter_example <- tibble::tibble(
  x = rep(0, 4),
  y = rep(0, 4)
)
jittered_plot_1 <- gg$ggplot(data = jitter_example, 
                             mapping = gg$aes(x = x, y = y)) +
  gg$geom_point() +
  gg$coord_cartesian(xlim = c(-0.025, 0.025), ylim = c(-0.025, 0.025)) +
  gg$labs(title = "Regular scatterplot")
jittered_plot_2 <- gg$ggplot(data = jitter_example, 
                             mapping = gg$aes(x = x, y = y)) +
  gg$geom_jitter(width = 0.01, height = 0.01) +
  gg$coord_cartesian(xlim = c(-0.025, 0.025), ylim = c(-0.025, 0.025)) +
  gg$labs(title = "Jittered scatterplot")
jittered_plot_1 + jittered_plot_2
```

In the left-hand regular scatterplot, 
observe that the 4 points are superimposed on top of each other. 
While we know there are 4 values being plotted, 
this fact might not be apparent to others. 
In the right-hand jittered scatterplot, 
it is now plainly evident that this plot involves four points 
since each point is given a random "nudge."  

Keep in mind, however, that jittering is strictly a visualization tool; 
even after creating a jittered scatterplot, 
the original values saved in the data frame remain unchanged. 
\index{ggplot2!geom\_jitter()}

To create a jittered scatterplot, 
instead of using `geom_point()`, we use `geom_jitter()`. 
Observe how the following code is very similar to the code 
that created the scatterplot with overplotting in Subsection \@ref(geompoint), 
but with `geom_point()` \index{ggplot2!geom\_point()} 
replaced with `geom_jitter()`. 

```{r jitter, fig.cap="Arrival versus departure delays jittered scatterplot."}
gg$ggplot(data = df_alaska_flights, 
          mapping = gg$aes(x = dep_delay, y = arr_delay)) + 
  gg$geom_jitter(width = 30, height = 30)
```

In order to specify how much jitter to add, 
we adjusted the `width` and `height` arguments to `geom_jitter()`. 
This corresponds to how hard you'd like to shake the plot 
in horizontal x-axis units and vertical y-axis units, respectively. 
In this case, both axes are in minutes. 
How much jitter should we add using the `width` and `height` arguments? 
On the one hand, it is important to add just enough jitter 
to break any overlap in points, 
but on the other hand, not so much 
that we completely alter the original pattern in points.

As can be seen in the resulting Figure \@ref(fig:jitter), 
in this case jittering doesn't really provide much new insight. 
In this particular case, 
it can be argued that changing the transparency of the points 
by setting `alpha` proved more effective. 
When would it be better to use a jittered scatterplot? 
When would it be better to alter the points' transparency? 
There is no single right answer that applies to all situations. 
You need to make a subjective choice and own that choice. 
At the very least when confronted with overplotting, however, 
we suggest you make both types of plots 
and see which one better emphasizes the point you are trying to make. 


```{block lc-overplotting, type="learncheck", purl=FALSE}
\vspace{-0.15in}
**_Learning check_**
\vspace{-0.1in}
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
After viewing Figure \@ref(fig:alpha), 
give an approximate range of arrival delays that occur most frequently. 
How about departure delays?
Now compare Figure \@ref(fig:alpha) to Figure \@ref(fig:noalpha). 
How has that region changed compared to 
when you observed the same plot without `alpha = 0.2` set 
in Figure \@ref(fig:noalpha)?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What additional information does it give you 
by setting the `alpha` argument that a regular scatterplot cannot?


```{block, type="learncheck", purl=FALSE}
\vspace{-0.25in}
\vspace{-0.25in}
```

### Summary

Scatterplots display the relationship between two continuous variables. 
They are among the most commonly used plots 
because they can provide an immediate way to see the trend 
in one continuous variable versus another. 
However, if you try to create a scatterplot 
where either one of the two variables is not continuous, 
you might get strange results.  Be careful! 

With medium to large datasets, 
you may need to play around with the different modifications to scatterplots 
we saw such as changing the transparency/opacity of the points 
or by jittering the points. 
This tweaking is often a fun part of data visualization, 
since you'll have the chance to see different relationships 
emerge as you tinker with your plots.

## 5NG#2: Linegraphs {#linegraphs}

The next of the five named graphs are linegraphs. 
Linegraphs \index{linegraphs} show the relationship 
between two continuous variables when the variable on the x-axis, 
also called the *explanatory* variable\index{explanatory variable}, 
is of a sequential nature. 
In other words, there is an inherent ordering to the variable. 

The most common examples of linegraphs have some notion of time 
on the x-axis: hours, days, weeks, years, etc. 
Since time is sequential, 
we connect consecutive observations of the variable on the y-axis with a line. 
Linegraphs that have some notion of time on the x-axis are also called 
*time series* plots\index{time series plots}. 
Let's illustrate linegraphs using another dataset in the 
`nycflights13` \index{R packages!nycflights13} package: 
the `weather` data frame. 

Let's explore the `weather` data frame 
by running `View(nycflights13::weather)` or `glimpse(nycflights13::weather)`. 
Furthermore let's read the associated help file 
by running `?nycflights13::weather` to bring up the help file.

Observe that there is a variable called `temp` 
of hourly temperature recordings in *Fahrenheit* at weather stations 
near all three major airports in New York City: 
Newark (`origin` code `EWR`), 
John F. Kennedy International (`JFK`), 
and LaGuardia (`LGA`). 
However, instead of considering hourly temperatures for all days in 2013 
for all three airports, 
for simplicity let's only consider hourly temperatures 
at Newark airport for the first 15 days in January. 

Recall in Section \@ref(scatterplots), 
we used the `filter()` function to only choose the subset of rows of `flights` 
corresponding to Alaska Airlines flights. 
We similarly use `filter()` \index{dplyr!filter()} here, 
but by using the `&` operator 
we only choose the subset of rows of `weather` where the `origin` is `"EWR"`, 
the `month` is January, **and** the `day` is between `1` and `15`. 
Recall we performed a similar task in Section \@ref(scatterplots) 
when creating the `df_alaska_flights` data frame of only Alaska Airlines flights, 
a topic we'll explore more in Chapter \@ref(wrangling) on data wrangling.

Before applying `filter`, 
let's `import` the `weather` dataset from package `nycflights13`, 
so it can be referred to in the short form `df_weather`. 
Furthermore, let's convert Fahrenheit to Celsius. 

```{r}
# import the data "weather" and save it as "df_weather"
import::from(nycflights13, df_weather = weather)
# convert Fahrenheit to Celsius
df_weather$temp_c = (df_weather$temp - 32) * (5 / 9)
```

```{r}
# Reduce the data to the intended subset
early_january_weather <- df_weather %>% 
  dp$filter(origin == "EWR" & month == 1 & day <= 15)
```

```{block lc-early_january_weather, type="learncheck", purl=FALSE}
\vspace{-0.15in}
**_Learning check_**
\vspace{-0.1in}
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
Take a look at both the `df_weather` and 
`early_january_weather` data frames 
by running `View(df_weather)` and `View(early_january_weather)`. 
In what respect do these data frames differ?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
`View()` the `df_flights` data frame again. 
Why does the `time_hour` variable uniquely identify the hour of the measurement, 
whereas the `hour` variable does not? 

```{block, type="learncheck", purl=FALSE}
\vspace{-0.25in}
\vspace{-0.25in}
```


### Linegraphs via `geom_line` {#geomline}

Let's create a time series plot of the hourly temperatures 
saved in the `early_january_weather` data frame 
by using `geom_line()` to create a linegraph\index{ggplot2!geom\_line()}, 
instead of `geom_point()` which we used previously to create scatterplots:

```{r hourlytemp, fig.cap="Hourly temperature in Newark for January 1-15, 2013."}
gg$ggplot(data = early_january_weather, 
       mapping = gg$aes(x = time_hour, y = temp_c)) +
  gg$geom_line() + 
  gg$labs(y = "Temperature (Celsius)")
```

Similar to the `ggplot()` code that created the scatterplot of departure 
and arrival delays for Alaska Airlines flights in Figure \@ref(fig:noalpha), 
let's break down this code piece-by-piece in terms of the grammar of graphics:

Within the `ggplot()` function call, 
we specify two of the components of the grammar of graphics as arguments:

1. The `data` to be the `early_january_weather` data frame 
   by setting `data = early_january_weather`.
1. The `aes`thetic `mapping` by setting 
   `mapping = gg$aes(x = time_hour, y = temp_c)`. 
   Specifically, the variable `time_hour` maps to the `x` position aesthetic, 
   while the variable `temp_c` maps to the `y` position aesthetic.

We add a layer to the `ggplot()` function call using the `+` sign. 
The layer in question specifies the third component of the grammar:  
the `geom`etric object in question. 
In this case, the geometric object is a `line` set by specifying `geom_line()`. 

```{block lc-linegraph, type="learncheck", purl=FALSE}
\vspace{-0.15in}
**_Learning check_**
\vspace{-0.1in}
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
Plot a time series of `temp_c` for Newark Airport 
in the first 15 days of *June* 2013.

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
Why should linegraphs be avoided 
when there is not a clear ordering of the horizontal axis?

```{block, type="learncheck", purl=FALSE}
\vspace{-0.25in}
\vspace{-0.25in}
```


### Summary

Linegraphs, just like scatterplots, 
display the relationship between two continuous variables. 
However, it is preferred to use linegraphs over scatterplots 
when the variable on the x-axis (i.e., the explanatory variable) 
has an inherent ordering, such as some notion of time.  


## 5NG#3: Histograms {#histograms}


Let's consider the `temp_c` variable in the `df_weather` data frame once again, 
but unlike with the linegraphs in Section \@ref(linegraphs), 
let's say we *don't* care about its relationship with time, 
but rather we only care about how the values of `temp_c` **distribute**. 
In other words:

1. What are the smallest and largest values?
1. What is the "center" or "most typical" value?
1. How do the values spread out?
1. What are frequent and infrequent values?

One way to visualize this *distribution* \index{distribution} 
of this single variable `temp_c` is to plot them on a horizontal line: 

```{r temp-on-line, echo=F, warning=F, fig.height=0.8, fig.cap="Plot of hourly temperature recordings from NYC in 2013."}
gg$ggplot(data = df_weather, 
          mapping = gg$aes(x = temp_c, y = factor("A"))) +
  gg$geom_point() +
  gg$labs(x = "Temperature (Celsius)") + 
  gg$theme(
    axis.ticks.y = gg$element_blank(),
    axis.title.y = gg$element_blank(),
    axis.text.y = gg$element_blank()
  )
hist_title <- "Histogram of Hourly Temperature Recordings from NYC in 2013"
```

This gives us a general idea of how the values of `temp_c` distribute: 
observe that temperatures vary from around 
`r round(min(df_weather$temp_c, na.rm = TRUE), 0)`&deg;C 
up to `r round(max(df_weather$temp_c, na.rm = TRUE), 0)`&deg;C. 
Furthermore, there appear to be more recorded temperatures 
between 5&deg;C and 15&deg;C than outside this range. 
However, because of the high degree of overplotting in the points, 
it's hard to get a sense of exactly how many values 
are between say 5&deg;C and 10&deg;C.

What is commonly produced instead of Figure \@ref(fig:temp-on-line) 
is known as a \index{histograms} *histogram*.  
A histogram is a plot that visualizes the *distribution* 
of a continuous value as follows:

1. We first cut up the x-axis into a series of \index{histograms!bins} *bins*, 
   where each bin represents a range of values. 
1. For each bin, we count the number of observations 
   that fall in the range corresponding to that bin.
1. Then for each bin, we draw a bar whose height marks the corresponding count.

Let's drill-down on an example of a histogram, 
shown in Figure \@ref(fig:histogramexample).

```{r histogramexample, echo=FALSE, warning=F, fig.cap="Example histogram.", fig.height=2, purl=FALSE}
gg$ggplot(data = df_weather, 
          mapping = gg$aes(x = temp_c)) +
  gg$geom_histogram(binwidth = 5, boundary = 15, color = "white") + 
  gg$labs(x = "Temperature (Celsius)")
```

Let's focus only on temperatures between 0&deg;C and 20&deg;C for now. 
Observe that there are four bins of equal width between 0&deg;C and 20&deg;C. 
Thus we have four bins of width 5&deg;C each: 
one bin for the 0-5&deg;C range, 
another bin for the 5-10&deg;C range, 
and another bin for the 10-15&deg;C range. Since:

1. The bin for the 5-10&deg;C range has a height of around 4000. 
   In other words, around 4000 of the hourly temperature recordings 
   are between 5&deg;C and 10&deg;C.
1. The bin for the 10-15&deg;C range has a height of around 4500. 
   In other words, around 4500 of the hourly temperature recordings 
   are between 10&deg;C and 15&deg;C.

All nine bins spanning -10&deg;C to 35&deg;C on the x-axis 
have this interpretation.


### Histograms via `geom_histogram` {#geomhistogram}

Let's now present the `ggplot()` code to plot your first histogram! 
Unlike with scatterplots and linegraphs, 
there is now only one variable being mapped in `aes()`: 
the single continuous variable `temp_c`. 
The y-aesthetic of a histogram, the count of the observations in each bin, 
gets computed for you automatically. 
Furthermore, the geometric object layer is now a `geom_histogram()`. 
\index{ggplot2!geom\_histogram()} 
After running the following code, 
you'll see the histogram in Figure \@ref(fig:weather-histogram) 
as well as warning messages. We'll discuss the warning messages first. 

```{r weather-histogram, warning=TRUE, fig.cap="Histogram of hourly temperatures at three NYC airports.", fig.height=2.3}
gg$ggplot(data = df_weather, 
          mapping = gg$aes(x = temp_c)) +
  gg$geom_histogram() + 
  gg$labs(x = "Temperature (Celsius)")
```

The first warning message is telling us that the histogram was constructed using `bins = 30` for 30 equally spaced bins. This is known in computer programming as a default value; unless you override this default number of bins with a number you specify, R will choose 30 by default. We'll see in the next section how to change the number of bins to another value than the default.

The second message is telling us something similar to the warning message we received when we ran the code to create a scatterplot of departure and arrival delays for Alaska Airlines flights in Figure \@ref(fig:noalpha): that because one row has a missing `NA` value for `temp_c`, it was omitted from the histogram. R is just giving us a friendly heads up that this was the case. 

Now let's unpack the resulting histogram in Figure \@ref(fig:weather-histogram). 
Observe that values less than -5&deg;C as well as values above 35&deg;C 
are rather rare. 
However, because of the large number of bins, 
it's hard to get a sense for which range of temperatures is spanned by each bin; 
everything is one giant amorphous blob. 
So let's add white vertical borders demarcating the bins 
by adding a `color = "white"` argument to `geom_histogram()` 
and ignore the warning about setting the number of bins to a better value:

```{r weather-histogram-2, message=FALSE, fig.cap="Histogram of hourly temperatures at three NYC airports with white borders.", fig.height=3}
gg$ggplot(data = df_weather, 
          mapping = gg$aes(x = temp_c)) +
  gg$geom_histogram(color = "white") + 
  gg$labs(x = "Temperature (Celsius)")
```

We now have an easier time associating ranges of temperatures 
to each of the bins in Figure \@ref(fig:weather-histogram-2). 
We can also vary the color of the bars 
by setting the \index{ggplot2!fill} `fill` argument. 
For example, you can set the bin colors to be "blue steel" 
by setting `fill = "steelblue"`:

```{r eval=FALSE}
# Try on your own
gg$ggplot(data = df_weather, 
          mapping = gg$aes(x = temp_c)) +
  gg$geom_histogram(color = "white", fill = "steelblue") + 
  gg$labs(x = "Temperature (Celsius)")
```

If you're curious, run \index{colors()} `colors()` to see 
all `r colors() %>% length()` possible choice of colors in R!

### Adjusting the bins {#adjustbins}

Observe in Figure \@ref(fig:weather-histogram-2) that 
in the 0-10&deg;C range there appear to be 6 bins. 
Thus each bin has a width of 10 divided by 6, 
or  1.667&deg;C, which is not a very easily interpretable range to work with. 
Let's improve this by adjusting the number of bins 
in our histogram in one of two ways:

1. By adjusting the number of bins via the \index{geom\_histogram()!bins} 
   `bins` argument in `geom_histogram()`. 
1. By adjusting the width of the bins via the \index{geom\_histogram()!binwidth} 
   `binwidth` argument in `geom_histogram()`. 

Using the first method, 
we have the power to specify how many bins 
we would like to cut the x-axis up in. 
As mentioned in the previous section, 
the default number of bins is 30. 
We can override this default, to say 40 bins, as follows:

```{r eval=FALSE}
gg$ggplot(data = df_weather, 
          mapping = gg$aes(x = temp_c)) +
  gg$geom_histogram(bins = 40, color = "white") + 
  gg$labs(x = "Temperature (Celsius)")
```

Using the second method, 
instead of specifying the number of bins, 
we specify the width of the bins by using the `binwidth` argument 
in the `geom_histogram()` layer. 
For example, let's set the width of each bin to be 5&deg;C.

```{r eval=FALSE}
gg$ggplot(data = df_weather, 
          mapping = gg$aes(x = temp_c)) +
  gg$geom_histogram(binwidth = 5, color = "white") + 
  gg$labs(x = "Temperature (Celsius)")
```

We compare both resulting histograms side-by-side in Figure \@ref(fig:hist-bins). 

```{r hist-bins, message=FALSE, fig.cap= "Setting histogram bins in two ways.", echo=FALSE, purl=FALSE}
hist_1 <- gg$ggplot(data = df_weather, 
                    mapping = gg$aes(x = temp_c)) +
  gg$geom_histogram(bins = 40, color = "white") +
  gg$labs(x = "Temperature (Celsius)", title = "With 40 bins")
hist_2 <- gg$ggplot(data = df_weather, 
                    mapping = gg$aes(x = temp_c)) +
  gg$geom_histogram(binwidth = 5, color = "white") +
  gg$labs(x = "Temperature (Celsius)", title = "With binwidth = 5 degrees C")
hist_1 + hist_2
```

```{block lc-histogram, type="learncheck", purl=FALSE}
\vspace{-0.15in}
**_Learning check_**
\vspace{-0.1in}
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
Change the number of bins to 20 and redraw the plot. 
How is the plot different from when the number of bins is 30? 40?
What can you conclude about how to choose the number of bins?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
Would you classify the distribution of temperatures as symmetric 
or skewed in one direction or another?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What would you guess is the "center" value in this distribution?  
Why did you make that choice?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
Is this data spread out greatly from the center or is it close?  Why?

```{block, type="learncheck", purl=FALSE}
\vspace{-0.25in}
\vspace{-0.25in}
```



### Summary


Histograms, unlike scatterplots and linegraphs, 
present information on only a single continuous variable. 
Specifically, they are visualizations of the distribution 
of the continuous variable in question. 




## Facets {#facets}


Before continuing with the next of the 5NG, 
let's briefly introduce a new concept called *faceting*.  
Faceting is used when we'd like to split a particular visualization 
by the values of another variable. 
This will create multiple copies of the same type of plot 
with matching x and y axes, but whose content will differ. 

For example, suppose we were interested in looking at 
how the histogram of hourly temperature recordings at the three NYC airports 
we saw in Figure \@ref(fig:histogramexample) *differed in each month*. 
We could "split" this histogram by the 12 possible months in a given year. 
In other words, we would plot histograms of `temp_c` for each `month` separately. 
We do this by adding `facet_wrap(~ month)` layer. 
Note the `~` is a "tilde" and you can type it 
by pressing `shift` and the key next to the "1" key on most US keyboards. 
The tilde is required and you'll receive the error 
`Error in as.quoted(facets) : object 'month' not found` 
if you don't include it here. 

```{r facethistogram, fig.cap="Faceted histogram of hourly temperatures by month.", purl=FALSE, fig.height=3.3}
gg$ggplot(data = df_weather, 
          mapping = gg$aes(x = temp_c)) +
  gg$geom_histogram(binwidth = 5, color = "white") +
  gg$facet_wrap(~ month) + 
  gg$labs(x = "Temperature (Celsius)")
```

We can also specify the number of rows and columns in the grid 
by using the `nrow` and `ncol` arguments 
inside of \index{ggplot2!facet\_wrap()} `facet_wrap()`. 
For example, say we would like our faceted histogram 
to have 4 rows instead of 3. 
We simply add an `nrow = 4` argument to `facet_wrap(~ month)`.

```{r facethistogram2, fig.cap="Faceted histogram with 4 instead of 3 rows.", purl=FALSE, fig.height=3.3}
gg$ggplot(data = df_weather, 
          mapping = gg$aes(x = temp_c)) +
  gg$geom_histogram(binwidth = 5, color = "white") +
  gg$facet_wrap(~ month, nrow = 4) + 
  gg$labs(x = "Temperature (Celsius)")
```


Observe in both Figures \@ref(fig:facethistogram) 
and \@ref(fig:facethistogram2) that 
as we might expect in the Northern Hemisphere, 
temperatures tend to be higher in the summer months, 
while they tend to be lower in the winter. 

```{block lc-facet, type="learncheck", purl=FALSE}
\vspace{-0.15in}
**_Learning check_**
\vspace{-0.1in}
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What other things do you notice about this faceted plot? 
How does a faceted plot help us see relationships between two variables?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What do the numbers 1-12 correspond to in the plot? 
What about 0, 20, 40?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
For which types of datasets would faceted plots 
not work well? 
Give an example of such a dataset and 
describe some of its important characteristics. 

```{block, type="learncheck", purl=FALSE}
\vspace{-0.25in}
\vspace{-0.25in}
```

## 5NG#4: Boxplots {#boxplots}

```{r echo=FALSE, purl=FALSE}
n_nov <- df_weather %>%
  dp$filter(month == 11) %>%
  nrow()
min_nov <- df_weather %>%
  dp$filter(month == 11) %>%
  dp$pull(temp_c) %>%
  min(na.rm = TRUE) %>%
  round(0)
max_nov <- df_weather %>%
  dp$filter(month == 11) %>%
  dp$pull(temp_c) %>%
  max(na.rm = TRUE) %>%
  round(0)
quartiles <- df_weather %>%
  dp$filter(month == 11) %>%
  dp$pull(temp_c) %>%
  quantile(prob = c(0.25, 0.5, 0.75)) %>%
  round(0)
five_number_summary <- tibble::tibble(
  temp = c(min_nov, quartiles, max_nov)
)
```

Faceted histograms are one type of visualization 
used to compare the distribution of a continuous variable 
split by the values of another variable. 
Another type of visualization that achieves this same goal 
is a *side-by-side boxplot*. 
A boxplot \index{boxplots} is constructed from the information 
provided in the *five-number summary* of a continuous variable 
(see Appendix \@ref(appendix-stat-terms)).  

To keep things simple for now, 
let's only consider the `r n_nov` hourly temperature recordings 
for the month of November, 
each represented as a jittered point in Figure \@ref(fig:nov1). 

```{r nov1, echo=FALSE, fig.cap="November temperatures represented as jittered points.", out.width="50%", purl=FALSE}
base_plot <- df_weather %>%
  dp$filter(month == 11) %>%
  gg$ggplot(mapping = gg$aes(x = factor(month), y = temp_c)) +
  gg$labs(x = "November", y = "Temperature (Celsius)") + 
  gg$theme(axis.ticks.x=gg$element_blank(),
           axis.text.x=gg$element_blank(), 
  )
base_plot +
  gg$geom_jitter(width = 0.075, height = 0.5, alpha = 0.3) + 
  gg$theme(
           axis.title=gg$element_text(size=gg$rel(2)), 
           axis.text.y=gg$element_text(size=gg$rel(2))
           )
```

These `r n_nov` observations have the following *five-number summary*:

1. Minimum: `r min_nov`&deg;C
1. First quartile (25th percentile): `r quartiles[1]`&deg;C
1. Median (second quartile, 50th percentile): `r quartiles[2]`&deg;C
1. Third quartile (75th percentile): `r quartiles[3]`&deg;C
1. Maximum: `r max_nov`&deg;C

In the leftmost plot of Figure \@ref(fig:nov2), 
let's mark these 5 values with dashed horizontal lines 
on top of the `r n_nov` points. 
In the middle plot of Figure \@ref(fig:nov2) 
let's add the *boxplot*. 
In the rightmost plot of Figure \@ref(fig:nov2), 
let's remove the points and the dashed horizontal lines for clarity's sake.

```{r nov2, echo=FALSE, fig.cap="Building up a boxplot of November temperatures.", fig.height=3, purl=FALSE}
boxplot_1 <- base_plot +
  gg$geom_hline(data = five_number_summary, 
                mapping = gg$aes(yintercept = temp), linetype = "dashed") +
  gg$geom_jitter(width = 0.075, height = 0.5, alpha = 0.1) + 
  gg$labs(y = "Temperature (Celsius)")

boxplot_2 <- base_plot +
  gg$geom_boxplot() +
  gg$geom_hline(data = five_number_summary, 
                mapping = gg$aes(yintercept = temp), linetype = "dashed") +
  gg$geom_jitter(width = 0.075, height = 0.5, alpha = 0.1) + 
  gg$labs(y = "Temperature (Celsius)")
boxplot_3 <- base_plot +
  gg$geom_boxplot() + 
  gg$labs(y = "Temperature (Celsius)")

boxplot_1 + boxplot_2 + boxplot_3
```

What the boxplot does is visually summarize the 
`r df_weather %>% dp$filter(month == 11) %>% nrow()` points 
by cutting the `r n_nov` temperature recordings 
into *quartiles* at the dashed lines, 
where each quartile contains roughly 
`r n_nov` $\div$ 4 $\approx$ `r round(n_nov / 4)` observations. Thus

```{r echo=FALSE, purl=FALSE}
# This redundant code is used for dynamic non-static in-line text output purposes
quartile_1 <- quartiles[1] %>% round(3)
quartile_2 <- quartiles[2] %>% round(3)
quartile_3 <- quartiles[3] %>% round(3)
```
  
1. 25% of points fall below the bottom edge of the box, 
   which is the first quartile of `r quartile_1`&deg;C. 
   In other words, 25% of observations were below `r quartile_1`&deg;C.
1. 25% of points fall between the bottom edge of the box 
   and the solid middle line, 
   which is the median of `r quartile_2`&deg;C. 
   Thus, 25% of observations were between `r quartile_1`&deg;C 
   and `r quartile_2`&deg;C and 50% of observations 
   were below `r quartile_2`&deg;C.
1. 25% of points fall between the solid middle line 
   and the top edge of the box, 
   which is the third quartile of `r quartile_3`&deg;C. 
   It follows that 25% of observations were 
   between `r quartile_2`&deg;C and `r quartile_3`&deg;C 
   and 75% of observations were below `r quartile_3`&deg;C.
1. 25% of points fall above the top edge of the box. 
   In other words, 25% of observations were above `r quartile_3`&deg;C.
1. The middle 50% of points lie within the *interquartile range (IQR)* 
   \index{interquartile range (IQR)} between the first and third quartile. 
   Thus, the IQR for this example is 
   `r quartile_3` - `r quartile_1` = 
   `r (quartiles[3] - quartiles[1]) %>% round(3)`&deg;C. 
   The interquartile range is a measure of a continuous variable's *spread*.

Furthermore, in the rightmost plot of Figure \@ref(fig:nov2), 
we see the *whiskers* \index{boxplots!whiskers} of the boxplot. 
The whiskers stick out from either end of the box 
all the way to the minimum and maximum observed temperatures 
of `r min_nov`&deg;C and `r max_nov`&deg;C, respectively. 
However, the whiskers don't always extend to the smallest 
and largest observed values as they do here. 
They in fact extend no more than 1.5 $\times$ the interquartile range 
from either end of the box. 
In this case of the November temperatures, 
no more than 1.5 $\times$ `r (quartiles[3] - quartiles[1]) %>% round(3)`&deg;C 
= `r (1.5*(quartiles[3] - quartiles[1])) %>% round(3)`&deg;C 
from either end of the box. 
Any observed values outside this range 
get marked with points called *outliers*, which we'll see in the next section.

### Boxplots via `geom_boxplot` {#geomboxplot}

Let's now create a side-by-side boxplot \index{boxplots!side-by-side} 
of hourly temperatures split by the 12 months 
as we did previously with the faceted histograms. 
We do this by mapping the `month` variable to the x-position aesthetic, 
the `temp_c` variable to the y-position aesthetic, 
and by adding a `geom_boxplot()` layer:


```{r monthtempbox, fig.cap="Side-by-side boxplot of temperature split by month."}
gg$ggplot(data = df_weather, 
          mapping = gg$aes(x = factor(month, ordered=T), y = temp_c)) +
  gg$geom_boxplot() + 
  gg$labs(y = "Temperature (Celsius)")
```

The code above shares the same structure as what we have seen 
with scatterplots and linegraphs, 
with one distinguished feature: `x = factor(...)`. 
We will discuss this distinction in section \@ref(boxplot-factor). 
For now, let's focus on the main theme: boxplot. 

The resulting Figure \@ref(fig:monthtempbox) 
shows 12 separate "box and whiskers" plots 
similar to the rightmost plot of Figure \@ref(fig:nov2) 
of only November temperatures. 
Thus the different boxplots are shown "side-by-side."

* The "box" portions of the visualization represent the 1st quartile, 
  the median (the 2nd quartile), and the 3rd quartile.
* The height of each box (the value of the 3rd quartile 
  minus the value of the 1st quartile) is the interquartile range (IQR). 
  It is a measure of the spread of the middle 50% of values, 
  with longer boxes indicating more variability.
* The "whisker" portions of these plots extend out from the bottoms 
  and tops of the boxes and represent points less than the 25th percentile 
  and greater than the 75th percentiles, respectively. 
  They're set to extend out no more than $1.5 \times IQR$ units away 
  from either end of the boxes. 
  We say "no more than" because the ends of the whiskers 
  have to correspond to observed temperatures.  
  The length of these whiskers show 
  how the data outside the middle 50% of values vary, 
  with longer whiskers indicating more variability.
* The dots representing values falling outside the whiskers 
  are called \index{outliers} *outliers*. 
  These can be thought of as anomalous ("out-of-the-ordinary") values.

**It is important to keep in mind that the definition of an outlier 
is somewhat arbitrary and not absolute.** 
In this case, they are defined by the length of the whiskers, 
which are no more than $1.5 \times IQR$ units long for each boxplot. 
Looking at this side-by-side plot we can see, 
as expected, that summer months (6 through 8) have higher median temperatures 
as evidenced by the higher solid lines in the middle of the boxes. 
We can easily compare temperatures across months 
by drawing imaginary horizontal lines across the plot. 
Furthermore, the heights of the 12 boxes as quantified 
by the interquartile ranges are informative too; 
they tell us about variability, or spread, 
of temperatures recorded in a given month. 


```{block lc-boxplot, type="learncheck", purl=FALSE}
\vspace{-0.15in}
**_Learning check_**
\vspace{-0.1in}
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What does the dot at the bottom of the plot for May correspond to?  
Explain what might have occurred in May to produce this point.

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
Which months have the highest variability in temperature?  
What reasons can you give for this?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
Boxplots provide a simple way to identify outliers.  
Why may outliers be easier to identify 
when looking at a boxplot instead of a faceted histogram?

```{block, type="learncheck", purl=FALSE}
\vspace{-0.25in}
\vspace{-0.25in}
```

### Summary

Side-by-side boxplots provide us with a way to compare the distribution 
of a continuous variable, such as temperature, 
across multiple values of another categorical variable. 
One can see where the median falls across the different groups 
by comparing the solid lines in the center of the boxes. 

To study the spread of a continuous variable within one of the boxes, 
look at both the length of the box 
and also how far the whiskers extend from either end of the box. 
Outliers are even more easily identified when looking at a boxplot 
than when looking at a histogram as they are marked with distinct points.

## 5NG#5: Barplots {#geombar}

A boxplot is often used to visualize the distribution of continuous variables. 
whereas a **barplot** is used to visualize the distribution 
of a categorical variable. 
Take sex as a categorical variable for example. 
We can visualize the number of male vs. female studuents in this class. 
To do so, we begin by counting heads in each category of sex, 
namely male and female. 
These categories are also known as 
the \index{levels} *levels* of the categorical variable. 
And the counts of each level are known as \index{frequencies} *frequencies*. 
A barplot is sometimes referred to as a barchart. 

Barplots are easy to construct. 
One complication, however, is how your data is represented. 
Is the categorical variable of interest "pre-counted" or not? 
For example, run the following code that manually creates two data frames 
representing a collection of fruit: 3 apples and 2 oranges.

```{r}
fruits <- tibble::tibble(
  fruit = c("apple", "apple", "orange", "apple", "orange")
)
fruits_counted <- tibble::tibble(
  fruit = c("apple", "orange"),
  number = c(3, 2)
)
```

We see both the `fruits` and `fruits_counted` data frames 
represent the same collection of fruit. 
Whereas `fruits` just lists the fruit individually...

```{r fruits, echo=FALSE, purl=FALSE}
fruits
```

... `fruits_counted` has a variable `count` which represent the "pre-counted" values of each fruit. 

```{r fruitscounted, echo=FALSE, purl=FALSE}
fruits_counted
```

Depending on how your categorical data is represented, 
you'll need to add a different `geom`etric layer type to your `ggplot()` 
to create a barplot, as we now explore. 


### Barplots via `geom_bar` or `geom_col`

Let's generate barplots using these two different representations 
of the same basket of fruit: 3 apples and 2 oranges. 
Using the `fruits` data frame 
where all 5 fruits are listed individually in 5 rows, 
we map the `fruit` variable to the x-position aesthetic 
and add a \index{ggplot2!geom\_bar()} `geom_bar()` layer:

```{r geombar, fig.cap="Barplot when counts are not pre-counted."}
gg$ggplot(data = fruits, 
          mapping = gg$aes(x = fruit)) +
  gg$geom_bar()
```

However, using the `fruits_counted` data frame 
where the fruits have been "pre-counted", 
we once again map the `fruit` variable to the x-position aesthetic, 
but here we also map the `count` variable to the y-position aesthetic, 
and add a \index{ggplot2!geom\_col()} `geom_col()` layer instead.

```{r geomcol, fig.cap="Barplot when counts are pre-counted."}
gg$ggplot(data = fruits_counted, 
          mapping = gg$aes(x = fruit, y = number)) +
  gg$geom_col()
```

Compare the barplots in Figures \@ref(fig:geombar) and \@ref(fig:geomcol). 
They are identical because they reflect counts of the same five fruits. 
However, depending on how our categorical data is represented, 
either "pre-counted" or not, we must add a different `geom` layer. 
When the categorical variable whose distribution you want to visualize

* Is *not* pre-counted in your data frame, we use `geom_bar()`.
* Is pre-counted in your data frame, we use `geom_col()` 
  with the y-position aesthetic mapped to the variable that has the counts.

Let's now go back to the `df_flights` data frame in the `nycflights13` package 
and visualize the distribution of the categorical variable `carrier`. 
In other words, let's visualize the number of domestic flights 
out of New York City each airline company flew in 2013. 
Recall from Subsection \@ref(exploredataframes) 
when you first explored the `df_flights` data frame, 
you saw that each row corresponds to a flight. 
In other words, the `df_flights` data frame is more like the `fruits` data frame 
than the `fruits_counted` data frame 
because the flights have not been pre-counted by `carrier`. 
Thus we should use `geom_bar()` instead of `geom_col()` to create a barplot. 
Much like a `geom_histogram()`, 
there is only one variable in the `aes()` aesthetic mapping: 
the variable `carrier` gets mapped to the `x`-position. 

(ref:geombar) Number of flights departing NYC in 2013 by airline using geom_bar().

```{r flightsbar, fig.cap="(ref:geombar)"}
gg$ggplot(data = df_flights, 
          mapping = gg$aes(x = carrier)) +
  gg$geom_bar()
```

Observe in Figure \@ref(fig:flightsbar) that United Airlines (UA), 
JetBlue Airways (B6), and ExpressJet Airlines (EV) 
had the most flights depart NYC in 2013. 
If you don't know which airlines correspond to which carrier codes, 
then run `View(nycflights13::airlines)` 
to see a directory of airlines. 
For example, B6 is JetBlue Airways. 
Alternatively, say you had a data frame where the number of flights 
for each `carrier` was pre-counted as in Table \@ref(tab:flights-counted).

```{r flights-counted, message=FALSE, echo=FALSE, purl=FALSE}
flights_counted <- df_flights %>%
  dp$group_by(carrier) %>%
  dp$summarize(number = dp$n())
knitr::kable(flights_counted,
  digits = 3,
  caption = "Number of flights pre-counted for each carrier",
  booktabs = TRUE,
  longtable = TRUE,
  linesep = ""
) %>%
  kableExtra::kable_styling(
    font_size = 16
  )
```

In order to create a barplot visualizing the distribution 
of the categorical variable `carrier` in this case, 
we would now use `geom_col()` instead of `geom_bar()`, 
with an additional `y = number` in the aesthetic mapping 
on top of the `x = carrier`. 
The resulting barplot would be identical to Figure \@ref(fig:flightsbar).

```{block lc-barplot, type="learncheck", purl=FALSE}
\vspace{-0.15in}
**_Learning check_**
\vspace{-0.1in}
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
When would a barplot be more appropriate than a histogram? 
Hint: consider which type of variables are being visulized. 

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What is the difference between histograms and barplots?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
How many Envoy Air flights departed NYC in 2013?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What was the 7th highest airline for departed flights from NYC in 2013? 
How could we better present the table to get this answer quickly?

```{block, type='learncheck', purl=FALSE}
\vspace{-0.25in}
\vspace{-0.25in}
```

### Must avoid pie charts!

One of the most common plots used to visualize the distribution 
of categorical data is the \index{pie charts} pie chart. 
While they may seem harmless enough, 
pie charts actually present a problem 
in that humans are unable to judge angles well. 
As Naomi Robbins describes in her book, 
*Creating More Effective Graphs* [@robbins2013], 
we overestimate angles greater than 90 degrees 
and we underestimate angles less than 90 degrees. 
In other words, it is difficult for us to determine the relative size 
of one piece of the pie compared to another.  

Let's examine the same data used in our previous barplot 
of the number of flights departing NYC by airline 
in Figure \@ref(fig:flightsbar), 
but this time we will use a pie chart 
in Figure \@ref(fig:carrierpie). Try to answer the following questions:

* How much larger is the portion of the pie for 
  ExpressJet Airlines (`EV`)  compared to US Airways (`US`)? 
* What is the third largest carrier in terms of departing flights?
* How many carriers have fewer flights than United Airlines (`UA`)?

```{r carrierpie, echo=FALSE, fig.cap="The dreaded pie chart.", fig.height=4.8, purl=FALSE}
gg$ggplot(df_flights, 
       mapping = gg$aes(x = factor(1), fill = carrier)) +
  gg$geom_bar(width = 1) +
  gg$coord_polar(theta = "y") +
  gg$theme(
    axis.title.x = gg$element_blank(),
    axis.title.y = gg$element_blank(),
    axis.ticks = gg$element_blank(),
    axis.text.y = gg$element_blank(),
    axis.text.x = gg$element_blank(),
    panel.grid.major = gg$element_blank(),
    panel.grid.minor = gg$element_blank()
  ) +
  gg$guides(fill = gg$guide_legend(keywidth = 0.8, keyheight = 0.8))
```

While it is quite difficult to answer these questions 
when looking at the pie chart in Figure \@ref(fig:carrierpie), 
we can much more easily answer these questions using the barchart 
in Figure \@ref(fig:flightsbar). 
This is true since barplots present the information 
in a way such that comparisons between categories 
can be made with single horizontal lines, 
whereas pie charts present the information 
in a way such that comparisons must be made by 
\index{pie charts!problems with} comparing angles. 

```{block lc-pie-charts, type="learncheck", purl=FALSE}
\vspace{-0.15in}
**_Learning check_**
\vspace{-0.1in}
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** Why do you think people continue to use pie charts?

```{block, type="learncheck", purl=FALSE}
\vspace{-0.25in}
\vspace{-0.25in}
```

### Two categorical variables {#two-categ-barplot}

Barplots are a very common way to visualize the frequency 
of different categories, or levels, of a single categorical variable. 
Another use of barplots is to visualize the *joint* distribution 
of two categorical variables at the same time.  
Let's examine the *joint* distribution of outgoing domestic flights 
from NYC by `carrier` as well as `origin`. 
In other words, the number of flights 
for each `carrier` and `origin` combination. 

For example, the number of WestJet flights from `JFK`, 
the number of WestJet flights from `LGA`, 
the number of WestJet flights from `EWR`, 
the number of American Airlines flights from `JFK`, 
and so on. 
Recall the `ggplot()` code that created the barplot of `carrier` frequency 
in Figure \@ref(fig:flightsbar):

```{r eval=FALSE}
gg$ggplot(data = df_flights, 
          mapping = gg$aes(x = carrier)) + 
  gg$geom_bar()
```

We can now map the additional variable `origin` 
by adding a `fill = origin` inside the `aes()` aesthetic mapping.

```{r flights-stacked-bar, fig.cap="Stacked barplot of flight amount by carrier and origin."}
gg$ggplot(data = df_flights, 
          mapping = gg$aes(x = carrier, fill = origin)) +
  gg$geom_bar()
```

Figure \@ref(fig:flights-stacked-bar) is an example 
of a \index{barplot!stacked} *stacked barplot*. 
While simple to make, in certain aspects it is not ideal. 
For example, it is difficult to compare the heights 
of the different colors between the bars, 
corresponding to comparing the number of flights 
from each `origin` airport between the carriers.

Before we continue, let's address some common points of 
confusion among new R users. 
First, the `fill` aesthetic corresponds to the color used to fill the bars, 
while the `color` aesthetic corresponds to the color of the outline of the bars. 
This is identical to how we added color to our histogram 
in Subsection \@ref(geomhistogram): 
we set the outline of the bars to white by setting `color = "white"` 
and the colors of the bars to blue steel by setting `fill = "steelblue"`. 
Observe in Figure \@ref(fig:flights-stacked-bar-color) 
that mapping `origin` to `color` and not `fill` 
yields grey bars with different colored outlines.

```{r flights-stacked-bar-color, fig.cap="Stacked barplot with color aesthetic used instead of fill.", purl=FALSE}
gg$ggplot(data = df_flights, 
          mapping = gg$aes(x = carrier, color = origin)) +
  gg$geom_bar()
```


Second, note that `fill` is another aesthetic mapping 
much like `x`-position; 
thus we were careful to include it 
within the parentheses of the `aes()` mapping. 
The following code, where the `fill` aesthetic is specified 
outside the `aes()` mapping will yield an error. 
This is a fairly common error that new `ggplot` users make:

```{r eval=FALSE}
# Try on your own. Will yield an error. 
gg$ggplot(data = df_flights, 
          mapping = gg$aes(x = carrier), fill = origin) +
  gg$geom_bar()
```

An alternative to stacked barplots are 
\index{barplot!side-by-side} *side-by-side barplots*, 
also known as *dodged barplots*, 
as seen in Figure \@ref(fig:flights-dodged-bar-color). 
The code to create a side-by-side barplot 
is identical to the code to create a stacked barplot, 
but with a \index{ggplot2!position} `position = "dodge"` argument 
added to `geom_bar()`. 
In other words, we are overriding the default barplot type, 
which is a *stacked* barplot, 
and specifying it to be a side-by-side barplot instead.

```{r flights-dodged-bar-color, fig.cap="Side-by-side barplot comparing number of flights by carrier and origin.", purl=FALSE}
gg$ggplot(data = df_flights, 
          mapping = gg$aes(x = carrier, fill = origin)) +
  gg$geom_bar(position = "dodge")
```


Note the width of the bars for `AS`, `F9`, `FL`, `HA` and `YV` 
is different than the others. 
We can make one tweak to the `position` argument 
to get them to be the same size in terms of width 
as the other bars by using the more robust `position_dodge()` function.

```{r flights-dodged-bar-color-tweak, fig.cap="Side-by-side barplot comparing number of flights by carrier and origin (with formatting tweak).", purl=FALSE}
gg$ggplot(data = df_flights, 
          mapping = gg$aes(x = carrier, fill = origin)) +
  gg$geom_bar(position = gg$position_dodge(preserve = "single"))
```


Lastly, another type of barplot is a \index{barplot!faceted} *faceted barplot*. 
Recall in Section \@ref(facets) we visualized the distribution 
of hourly temperatures at the 3 NYC airports *split* by month using facets. 
We apply the same principle to our barplot 
visualizing the frequency of `carrier` split by `origin`: 
instead of mapping `origin` to `fill` 
we include it as the variable to create small multiples of the plot 
across the levels of `origin`.

```{r facet-bar-vert, fig.cap="Faceted barplot comparing the number of flights by carrier and origin.", purl=FALSE}
gg$ggplot(data = df_flights, 
          mapping = gg$aes(x = carrier)) +
  gg$geom_bar() +
  gg$facet_wrap(~ origin, ncol = 1)
```


```{block lc-barplot-stacked, type="learncheck", purl=FALSE}
\vspace{-0.15in}
**_Learning check_**
\vspace{-0.1in}
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What kinds of questions are not easily answered by looking at Figure \@ref(fig:flights-stacked-bar)?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What can you say, if anything, about the relationship 
between airline and airport in NYC in 2013 
in regards to the number of departing flights?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
Why might the side-by-side barplot be preferable 
to a stacked barplot in this case?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What are the disadvantages of using a dodged barplot, in general?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
Why is the faceted barplot preferred 
to the side-by-side and stacked barplots in this case?

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What information about the different carriers 
at different airports is more easily seen in the faceted barplot?

```{block, type="learncheck", purl=FALSE}
\vspace{-0.25in}
\vspace{-0.25in}
```

### Summary

Barplots are a common way of displaying the distribution 
of a categorical variable, 
or in other words the frequency with which the different categories 
(also called *levels*) occur.  
They are easy to understand and make it easy to make comparisons across levels. 
Furthermore, when trying to visualize the relationship 
of two categorical variables, 
you have many options: stacked barplots, side-by-side barplots, 
and faceted barplots. 
Depending on what aspect of the relationship you are trying to emphasize, 
you will need to make a choice 
between these three types of barplots and own that choice.

## Conclusion {#data-vis-conclusion}

### Summary table

Let's recap all five of the five named graphs (5NG) 
\index{five named graphs} in Table \@ref(tab:viz-summary-table) 
summarizing their differences. 
Using these 5NG, 
you'll be able to visualize the distributions and relationships 
of variables contained in a wide array of datasets. 
This will be even more the case as we start to map more variables 
to more of each `geom`etric object's `aes`thetic attribute options, 
further unlocking the awesome power of the `ggplot2` package.

```{r viz-summary-table, echo=FALSE, message=FALSE, purl=FALSE}
# The following Google Doc is published to CSV and loaded using read_csv():
# https://docs.google.com/spreadsheets/d/1vzqlFiT6qm5wzy_L_0nL7EWAd6jiUZmLSCFhDhztDSg/edit#gid=0


if (!file.exists(here::here(
                            "rds", 
                            "summary_table_ch2.rds"))) {
  summary_table_ch2 <-
    "https://docs.google.com/spreadsheets/d/e/2PACX-1vRGaUW6EMIGPhg2V7CahoSdVi_JCcESFRYV5tov6bjcwOcn7DZDzfpZgrvjfFG6PV57gcJYIrwl_Q2c/pub?gid=0&single=true&output=csv" %>%
    readr::read_csv(na = "")
  summary_table_ch2$Notes[2] <- "Used when there is a sequential order to x-variable, e.g., time"
  summary_table_ch2$Notes[3] <- "Facetted histograms show the distribution of 1 continuous variable split by the values of another variable"
  summary_table_ch2$Shows[1] <- "Relationship between 2 continuous variables"
  summary_table_ch2$Shows[2] <- "Relationship between 2 continuous variables"
  summary_table_ch2$Shows[3] <- "Distribution of 1 continuous variable"
  summary_table_ch2$Shows[4] <- "Distribution of 1 continuous variable split by the values of another variable"
  readr::write_rds(summary_table_ch2, here::here("rds", 
                                                 "summary_table_ch2.rds"))
} else {
  summary_table_ch2 <- readr::read_rds(here::here("rds", 
                                                  "summary_table_ch2.rds"))
}

summary_table_ch2 %>% 
  dp$rename(` ` = `X1`) %>% 
  knitr::kable(
    caption = "Summary of Five Named Graphs",
    booktabs = TRUE,
    format = "html"
    )
```


### Function argument specification

Let's go over some important points about specifying the arguments 
(i.e., inputs) to functions. 
Run the following two segments of code:

```{r eval=FALSE}
# Segment 1:
gg$ggplot(data = df_flights, 
          mapping = gg$aes(x = carrier)) +
  gg$geom_bar()

# Segment 2:
gg$ggplot(df_flights, 
          gg$aes(x = carrier)) +
  gg$geom_bar()
```

You'll notice that both code segments create the same barplot, 
even though in the second segment 
we omitted the `data = ` and `mapping = ` code argument names. 
This is because the `ggplot()` function by default assumes that 
the `data` argument comes first 
and the `mapping` argument comes second. \index{functions!argument order} 
As long as you specify the data frame in question first 
and the `aes()` mapping second, 
you can omit the explicit statement of the argument names 
`data = ` and `mapping = `. 

Going forward for the rest of this book, 
all `ggplot()` code will be like the second segment: 
with the `data = ` and `mapping = ` explicit naming of the argument 
omitted with the default ordering of arguments respected. 
We'll do this for brevity's sake; 
it's common to see this style when reviewing other R users' code.



## Additional resources

### `factor()` in boxplot {#boxplot-factor}

In section \@ref(geomboxplot), we learned how to make 
side-by-side boxplots with the following code: 

```{r eval=F}
gg$ggplot(data = df_weather, 
          mapping = gg$aes(x = factor(month, ordered=T), y = temp_c)) +
  gg$geom_boxplot()
```

Structure-wise, 
this code resembles what we used to create scatterplots and linegraphs, 
which I reproduce below for you to compare:

```{r eval=FALSE}
# scatterplot
gg$ggplot(data = df_alaska_flights, 
          mapping = gg$aes(x = dep_delay, y = arr_delay)) + 
  gg$geom_point() + 
```

```{r eval=F}
# linegraph
gg$ggplot(data = early_january_weather, 
       mapping = gg$aes(x = time_hour, y = temp_c)) +
  gg$geom_line()
```

There is one distinguished feature in the boxplot code though: 

```
mapping = gg$aes(x = factor(month, ordered=T)) ...
```

Instead of `x = month`, we used `x = factor(month, ordered=T)`. 
Naturally, you will ask what we can achieve with `factor()`? 
And why do we need to apply `factor()` on the `month` variable 
for the boxplot? 


`factor()` is a base R function. 
It is typically applied to variables that are inherently categorical in nature, 
such as employment status, or days of a week. 
These variables share a common feature: 
they all have a countable number of *possible* values, 
rather than having a countless number of *possible* values. 
Take days of a week versus the height of a person for example. 
There are 7 possible values for the former: Monday, Tuesday, 
Wednesday, Thursday, Friday, Saturday and Sunday. 
In contrast, there is an infinite number of *possible* values 
for a person's height: 
1.6 m, 1.61 m, 1.611 m, 1.6111 m, ... You get the idea. 
Variables that have a *finite* number of *possible* values are often called 
**categorical** or **discrete** variables, 
whereas those that have an *infinite* number of *possible* values 
are called **continuous** variables. 

Note the italicized word *possible* in the previous paragraph. 
When judging whether a variable is categorical or continuous, 
it is important to think what its *possible* values are, 
rather than what its actual values are in a sample. 
A `height` variable may only have a finite number of *actual* values 
in a sample of 10 people: 1.58 m, 1.63 m, 1.79 m, 1.67 m, 1.72 m, 1.65 m, 
1.74 m, 1.82 m, 1.79 m, 1.81 m. 
However, as we have illustrated before, 
the possibilities are infinite. 
Therefore, height would be considered a continuous variable. 

Let's bring our attention back to the boxplot. 

`x = factor(month, ordered=T)`

In this example, 
`month` has exactly 12 possible values: "`1`" through "`12`". 
By applying `factor()` to the variable `month`, 
we are giving `R` explicit instruction to treat `month` as a categorical variable, 
with 12 levels. 
In addition, by setting `ordered` to `T` (short for `TRUE`), 
we are practically telling `R` to sort the 12 levels of `month`, 
with `1` being the first and `12` the last, 
matching the natural order of twelve months in a calendar year. 

What happens if you do not apply `factor()` to `month`, 
and just use `month` as is for the boxplot? 
Let's try it!

```{r badbox, warning=T, fig.cap="Invalid boxplot specification."}
gg$ggplot(data = df_weather, 
          mapping = gg$aes(x = month, y = temp_c)) +
  gg$geom_boxplot() + 
  gg$labs(y = "Temperature (Celsius)")
```

Observe in Figure \@ref(fig:badbox) that this plot 
does not look like Figure \@ref(fig:monthtempbox) at all. 
Unlike Figure \@ref(fig:monthtempbox), 
it does not provide information about temperature separated by month. 
The first warning message clues us in as to why. 
It is telling us that we have a "continuous" variable on the x-axis. 
But you may protest: 
is it not obvious that the `month` variable is a categorical one? 
After all, there are only "`1`" through "`12`", tweleve actual values 
in the data `df_weather`. 
Yes, as an adult English speaker, 
it is very obivous that `month` is a categorical variable, 
evidenced by its values only spanning the twelve integers 1 to 12. 
However, `R` is not an adult English speaker. 
If anything, it is more like an utterly stubborn polylingual rule stickler. 
It demands explicit instruction from us, the superior human users, 
on how to interpret a variable. 
Until we explicitly tell `R` that `month` is a categorical variable 
with ordered levels 1 through 12, 
it will treat `month` as a continuous one, 
with *possible* values of 1, 2, 3, ..., 10, 11, 12, 13, 14, ...

So we have seen what happens if we omit the `factor(..)` in the boxplot. 
`ggplot`'s boxplots require a categorical variable to be mapped 
to the x-position aesthetic. 
When the requirement is not met, 
`ggplot` throws out the unsatisfying `month`, 
gives a warning, 
and tries to only work with the y-axis, `temp_c`, 
which results in Figure \@ref(fig:badbox). 

The second warning message is equivalent to the warning message we have seen 
when plotting a histogram of hourly temperatures 
in Figure \@ref(fig:weather-histogram): 
that one of the rows has missing value `NA` in `temp_c` and/or `month`. 
To confirm: 

```{r eval=F}
dp$filter(df_weather, is.na(temp_c)|is.na(month))
```

```{r missing-data-2, echo=F}
dp$filter(df_weather, is.na(temp_c)|is.na(month)) %>% 
  knitr::kable(caption = "One row with missing data in `temp_c` and/or `month`") %>% 
	kableExtra::kable_styling(bootstrap_options = 
                            c("striped", "hover", "condensed", "responsive"), 
                          font_size = 14) %>% 
  kableExtra::column_spec(c(3,16), bold=T, color="white", background="#D7261E") %>% 
  kableExtra::scroll_box(width = "100%")
```

Sure enough, this row has a missing value in `temp_c`. 
(Place your mouse cursor inside the table and scroll right 
to see more columns.) 


```{block lc-factor, type="learncheck", purl=FALSE}
\vspace{-0.15in}
**_Learning check_**
\vspace{-0.1in}
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What happens when you completely drop the `x` axis in the boxplot? 

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
We looked at the distribution of the continuous variable `temp_c` 
split by the variable `month` that we converted 
using the `factor()` function in order to make a side-by-side boxplot. 
Why would a boxplot of `temp_c` split by the continuous variable `pressure` 
similarly converted to a categorical variable 
using the `factor()` not be informative?


```{block, type="learncheck", purl=FALSE}
\vspace{-0.25in}
\vspace{-0.25in}
```


### Common problems {#common-problems}

This section is from <https://r4ds.had.co.nz/data-visualisation.html#common-problems>

As you start to run R code, you're likely to run into problems. 
Don't worry --- it happens to everyone. 
I have been writing R code for years, 
and every day I still write code that doesn't work! 

Start by carefully comparing the code that you're running 
to the code in the book. 
R is extremely picky, and a misplaced character can make all the difference. 
Make sure that every `(` is matched with a `)` 
and every `"` is paired with another `"`. 
Sometimes you'll run the code and nothing happens. 
Check the left-hand of your console: if it's a `+`, 
it means that R doesn't think you've typed a complete expression 
and it's waiting for you to finish it. 
In this case, it's usually easy to start from scratch again 
by pressing ESCAPE to abort processing the current command.

One common problem when creating ggplot2 graphics 
is to put the `+` in the wrong place: 
it has to come at the end of the line, not the start. 
In other words, make sure you haven't accidentally written code like this:

```{r eval=F, tidy=F}
# Try by yourself, will encounter an error
ggplot2::ggplot(data = mpg) 
+ ggplot::geom_point(
                     mapping = ggplot2::aes(x = displ, y = hwy))
```

If you're still stuck, try the help (see section \@ref(help)). 
Don't worry if the help doesn't seem that helpful - 
instead, skip down to the examples and 
look for code that matches what you're trying to do.

If that doesn't help, carefully read the error message. 
Sometimes the answer will be buried there! 
But when you're new to R, the answer might be in the error message 
but you don't yet know how to understand it. 
Another great tool is Google: try googling the error message, 
as it's likely someone else has had the same problem, and has gotten help online.



### What's to come {#whats-to-come-3}

Recall in Figure \@ref(fig:noalpha) in Section \@ref(scatterplots) 
we visualized the relationship between departure delay and arrival delay 
for Alaska Airlines flights. 
This necessitated paring down the `df_flights` data frame 
to a new data frame `alaska_flights` 
consisting of only `carrier == AS` flights first:

```{r eval=FALSE}
alaska_flights <- df_flights %>% 
  dp$filter(carrier == "AS")

gg$ggplot(data = alaska_flights, 
          mapping = gg$aes(x = dep_delay, y = arr_delay)) + 
  gg$geom_point()
```

Furthermore recall in Figure \@ref(fig:hourlytemp) in Section \@ref(linegraphs) 
we visualized hourly temperature recordings at Newark airport 
only for the first 15 days of January 2013. 
This necessitated paring down the `df_weather` data frame 
to a new data frame `early_january_weather` 
consisting of hourly temperature recordings only for `origin == "EWR"`, 
`month == 1`, and day less than or equal to `15` first:

```{r eval=FALSE}
early_january_weather <- df_weather %>% 
  dp$filter(origin == "EWR" & month == 1 & day <= 15)

gg$ggplot(data = early_january_weather, 
          mapping = gg$aes(x = time_hour, y = temp_c)) +
  gg$geom_line()
```

These two code segments were a preview of Chapter \@ref(wrangling) 
on data wrangling using the `dplyr` package. 
Data wrangling is the process of transforming 
and modifying existing data with the intent of making it more appropriate 
for analysis purposes. 
For example, these two code segments used the `filter()` function 
to create new data frames (`alaska_flights` and `early_january_weather`) 
by choosing only a subset of rows of existing data frames 
(`flights` and `weather`). 
In the next chapter, we'll formally introduce the `filter()` 
and other data wrangling functions 
as well as the *pipe operator* `%>%` 
which allows you to combine multiple data wrangling actions 
into a single sequential *chain* of actions. 
On to Chapter \@ref(wrangling) on data wrangling!


