# Data Visualization {#viz}

```{r setup-viz, include=FALSE, purl=FALSE}
# take from the source
```


```{r load-package, message=FALSE, warning=FALSE}
#
# Install packages
#
# Install xfun so that I can use xfun::pkg_load2

cran_primary <- c("ggplot2", 
								 "dplyr", 
								 "nycflights13")
cran_secondary <- c("gapminder", 
									 "readr", 
									 "patchwork", 
									 "scales", 
									 "stringr"
									 )

if (length(cran_primary) != 0) xf$pkg_load2(cran_primary)
xf$pkg_load2(cran_primary, cran_secondary)
gg <- loadNamespace('ggplot2')
dp <- loadNamespace('dplyr')

import::from(magrittr, '%>%', '%T>%')
```


## Some core concepnts

layer

mapping


### An example

The joy of stats
[Hans Rosling](https://www.youtube.com/watch?v=jbkSRLYSojo)

Let's try to replicate part of the graphics. 
Health (life-exp) as a function of wealth (income) 
across different contries.

```{r, echo=FALSE, purl=FALSE}

```


### The grammar of graphics

graphic *layer* vs. linguistic element






### Components of the grammar

a graphic consists of 
data
geometric object (geom for short)
aesthetic attributes (aes for short)

