# Data Importing and "Tidy" Data {#tidy}

```{r setup_tidy, include=FALSE, purl=FALSE}
# Used to define Learning Check numbers:
chap <- 4
lc <- 0

# Set output digit precision
options(scipen = 99, digits = 3)

# Set random number generator see value for replicable pseudorandomness.
set.seed(76)
```

In Subsection \@ref(programming-concepts), 
we introduced the concept of a \index{data frame} data frame in R: 
a rectangular spreadsheet-like representation of data 
where the rows correspond to observations and the columns 
correspond to variables describing each observation.  
In Section \@ref(nycflights13), we started exploring our first data frame: 
the `flights` data frame included in the `nycflights13` package. 
In Chapter \@ref(viz), we created visualizations based on the data 
included in `flights` and other data frames such as `weather`. 
In Chapter \@ref(wrangling), we learned how to take existing data frames 
and transform/modify them to suit our ends. 

In this final chapter of the "Data Science with `tidyverse`" portion 
of the book, 
we extend some of these ideas 
by discussing a type of data formatting called "tidy" data. 
You will see that having data stored in "tidy" format 
is about more than just what the everyday definition 
of the term "tidy" might suggest: 
having your data "neatly organized." 
Instead, we define the term "tidy" as it's used by data scientists who use R, 
outlining a set of rules by which data is saved.

Knowledge of this type of data formatting 
was not necessary for our treatment of data visualization 
in Chapter \@ref(viz) and data wrangling in Chapter \@ref(wrangling). 
This is because all the data used were already in "tidy" format. 
In this chapter, we'll now see that this format 
is essential to using the tools we covered up until now. 
Furthermore, it will also be useful for all subsequent chapters 
in this book when we cover regression and statistical inference. 
First, however, we'll show you how to import spreadsheet data in R.

### Needed packages {-#tidy-packages}

The following code gives us access to the demo data, 
as well as the to some tools for us to interact with the data. 

```{r load-package, eval=F}
# Install xfun so that I can use xfun::pkg_load2
if (!requireNamespace('xfun')) install.packages('xfun')
xf <- loadNamespace('xfun')

cran_primary <- c(
                  "dplyr", 
                  "fivethirtyeight", 
                  "ggplot2", 
                  "here", 
                  "nycflights13", 
                  "readr", 
                  "readxl", 
                  "tidyr"
)

if (length(cran_primary) != 0) xf$pkg_load2(cran_primary)

import::from(magrittr, "%>%")
gg <- import::from(ggplot2, .all=TRUE, .into={new.env()})
dp <- import::from(dplyr, .all=TRUE, .into={new.env()})

import::from(nycflights13, 
df_airlines = airlines, 
df_airports = airports, 
df_flights = flights, 
df_planes = planes, 
df_weather = weather
)
```

```{r import-pkg, echo=F, message=FALSE, warning=FALSE}
import::from(magrittr, "%>%")
gg <- import::from(ggplot2, .all=TRUE, .into={new.env()})
dp <- import::from(dplyr, .all=TRUE, .into={new.env()})

import::from(nycflights13, 
df_airlines = airlines, 
df_airports = airports, 
df_flights = flights, 
df_planes = planes, 
df_weather = weather
)
```

## Importing data {#csv}

Up to this point, 
we've almost entirely used data stored inside of an R package. 
Say instead you have your own data saved on your computer or somewhere online. 
How can you analyze this data in R? 
Spreadsheet data is often saved in one of the following three formats:

First, a *Comma Separated Values* `.csv` \index{CSV file} file. 
You can think of a `.csv` file as a bare-bones spreadsheet where:

* Each line in the file corresponds to one row of data/one observation.
* Values for each line are separated with commas. 
  The values of different variables are separated by commas in each row.
* The first line is often, but not always, 
  a *header* row indicating the names of the columns/variables.

Second, an Excel `.xlsx` spreadsheet file. 
This format is based on Microsoft's proprietary Excel software. 
As opposed to bare-bones `.csv` files, 
`.xlsx` Excel files contain a lot of meta-data\index{meta-data} 
(data about data). 
Recall we saw a previous example of meta-data in Section \@ref(groupby) 
when adding "group structure" meta-data to a data frame 
by using the `group_by()` verb. 
Some examples of Excel spreadsheet meta-data 
include the use of bold and italic fonts, colored cells, 
different column widths, and formula macros.

Third, a [Google Sheets](https://www.google.com/sheets/about/) file, 
which is a "cloud" or online-based way to work with a spreadsheet. 
To import Google Sheets data in R, 
you can use a package called 
[`googlesheets`](https://cran.r-project.org/web/packages/googlesheets/vignettes/basic-usage.html), 
a method we will leave for more advanced books. 
Google Sheets also allows you to download your data 
in both comma separated values `.csv` and Excel `.xlsx` formats. 
Next we cover a method for importing `.csv` and `.xlsx` spreadsheet data in R. 

### `.csv` file

First, let's import a Comma Separated Values `.csv` file 
that exists on the internet. 
The `.csv` file `dem_score.csv` contains ratings of the level of democracy 
in different countries spanning 1952 to 1992 and 
is accessible at <https://moderndive.com/data/dem_score.csv>. 
Let's use the `read_csv()` function from the `readr` 
\index{R packages!readr!read\_csv()} [@R-readr] package 
to read it off the web, import it into R, 
and save it in a data frame called `dem_score`.

```{r message=FALSE, eval=FALSE}
dem_score <- readr::read_csv("https://moderndive.com/data/dem_score.csv")
dem_score
```
```{r message=FALSE, echo=FALSE, purl=FALSE}
dem_score <- readr::read_csv(here::here("data", "dem_score.csv"))
dem_score
```

In this `dem_score` data frame, 
the minimum value of `-10` corresponds to a highly autocratic nation, 
whereas a value of `10` corresponds to a highly democratic nation. 
Note the backticks "\`" that surround the different variable names.  
Variable names in R by default are not allowed to start with a number 
nor include spaces, 
but for aesthetic purposes, it is occasionally desirable to have those "illegal" 
variable names in a formated table. 
We can achieve this by surrounding the "illegal" names with backticks. 
We'll revisit the `dem_score` data frame 
in a case study in the upcoming Section \@ref(case-study-tidy).

Note that the `read_csv()` function included in the `readr` package 
is different than the `read.csv()` function in base R, 
which comes installed with R. 
While the difference in the names might seem trivial (a `_` instead of a `.`), 
the `read_csv()` function is, in our opinion, easier to use 
because it can more easily read data off the web 
and generally imports data at a much faster speed. 
Furthermore, the `read_csv()` function included in the `readr` 
saves data frames as `tibbles` by default. 

### `.xlsx` file

Next, let's import an Excel file, `dem_score.xlsx`. 
This file is identical to `dem_score.csv` we just imported, 
except that it has a different file extension. 
To read in an excel file, we need to use `read_excel()` from package`readxl`. 
Unlike `read_csv()` from `readr`, 
`read_excel()` cannot directly read in a file from the internet. 
Let's download the file first. 

If you are windows user:

```{r eval=F}
data_url <- "https://moderndive.com/data/dem_score.xlsx"
download.file(url=data_url, destfile=here::here("data", "dem_score.xlsx"), mode="wb")
```

If you are mac or linux user:

```{r eval=F}
data_url <- "https://moderndive.com/data/dem_score.xlsx"
xfun::download_file(url=data_url, output = here::here("data", "dem_score.xlsx"))
```

Now `dem_score.xlsx` should be present in the "data" foler inside your project. 
Let's import this data into R now. 

```{r eval=F}
dem_score <- readxl::read_excel(here::here("data", "dem_score.xlsx"))
dem_score
```


## "Tidy" data {#tidy-data-ex}

Let's now switch gears and learn about the concept of "tidy" data format 
with a motivating example from the `fivethirtyeight` package. 
The `fivethirtyeight` package [@R-fivethirtyeight] provides access 
to the datasets used in many articles published 
by the data journalism website, 
[FiveThirtyEight.com](https://fivethirtyeight.com/). 
For a complete list of all `r nrow(data(package = "fivethirtyeight")[[3]])` 
datasets included in the `fivethirtyeight` package, 
check out the package webpage by going to: 
<https://fivethirtyeight-r.netlify.app/articles/fivethirtyeight.html>.
\index{R packages!fivethirtyeight}

Let's focus our attention on the `drinks` data frame 
and look at its first 5 rows:

```{r}
import::from(fivethirtyeight, df_drinks = drinks)
head(df_drinks)
```

After reading the help file by running `?fivethirtyeight::drinks`, 
you'll see that `df_drinks` is a data frame 
containing results from a survey of the average number of servings of beer, 
spirits, and wine consumed in `r df_drinks %>% nrow()` countries. 
This data was originally reported on FiveThirtyEight.com 
in Mona Chalabi's article: 
["Dear Mona Followup: Where Do People Drink The Most Beer, Wine And Spirits?"](https://fivethirtyeight.com/features/dear-mona-followup-where-do-people-drink-the-most-beer-wine-and-spirits/).

Let's apply some of the data wrangling verbs we learned 
in Chapter \@ref(wrangling) on the `df_drinks` data frame:

1. `filter()` the `df_drinks` data frame to only consider 4 countries: 
   the United States, China, Italy, and Saudi Arabia, *then*
1. `select()` all columns except `total_litres_of_pure_alcohol` 
   by using the `-` sign, *then*
1. `rename()` the variables `beer_servings`, `spirit_servings`, 
   and `wine_servings` to `beer`, `spirit`, and `wine`, respectively.

and save the resulting data frame in `drinks_smaller`:

```{r}
drinks_smaller <- df_drinks %>% 
  dp$filter(country %in% c("USA", "China", "Italy", "Saudi Arabia")) %>% 
  dp$select(-total_litres_of_pure_alcohol) %>% 
  dp$rename(beer = beer_servings, spirit = spirit_servings, wine = wine_servings)
drinks_smaller
```

Let's now ask ourselves a question: 
"Using the `drinks_smaller` data frame, 
how would we create the side-by-side barplot 
in Figure \@ref(fig:drinks-smaller)?". 
Recall we saw barplots displaying two categorical variables 
in Subsection \@ref(two-categ-barplot).

```{r drinks-smaller, fig.cap="Comparing alcohol consumption in 4 countries.", echo=FALSE, purl=FALSE}
drinks_smaller_tidy <- drinks_smaller %>%
  tidyr::pivot_longer(names_to = "type", 
                      values_to = "servings", 
                      cols = -country)
drinks_smaller_tidy_plot <- gg$ggplot(
  drinks_smaller_tidy,
  gg$aes(x = country, y = servings, fill = type)
) +
  gg$geom_col(position = "dodge") +
  gg$labs(x = "country", y = "servings")
drinks_smaller_tidy_plot
```

Let's break down the grammar of graphics we introduced 
in Section \@ref(grammarofgraphics):

1. The categorical variable `country` with four levels 
   (China, Italy, Saudi Arabia, USA) 
   would have to be mapped to the `x`-position of the bars.
1. The numerical variable `servings` would have to be mapped 
   to the `y`-position of the bars (the height of the bars).
1. The categorical variable `type` with three levels (beer, spirit, wine) 
   would have to be mapped to the `fill` color of the bars.

Observe that `drinks_smaller` has three separate variables 
`beer`, `spirit`, and `wine`.  
In order to use the `ggplot()` function to recreate the barplot 
in Figure \@ref(fig:drinks-smaller) however, 
we need a *single variable* `type` with three possible values: 
`beer`, `spirit`, and `wine`.  
We could then map this `type` variable to the `fill` aesthetic of our plot.  
In other words, to recreate the barplot in Figure \@ref(fig:drinks-smaller), 
our data frame would have to look like this:

```{r echo=T, purl=FALSE}
drinks_smaller_tidy
```

```{r echo=FALSE, purl=FALSE}
# This redundant code is used for dynamic non-static in-line text output purposes
n_row_drinks <- drinks_smaller_tidy %>% nrow()
n_alcohol_types <- drinks_smaller_tidy %>%
  dp$select(type) %>%
  dp$n_distinct()
n_countries <- drinks_smaller_tidy %>%
  dp$select(country) %>%
  dp$n_distinct()
```

Observe that while `drinks_smaller` and `drinks_smaller_tidy` 
are both rectangular in shape 
and contain the same `r n_row_drinks` numerical values 
(`r n_alcohol_types` alcohol types by `r n_countries` countries), 
they are formatted differently. 
`drinks_smaller` is formatted in what's known as \index{wide data format} 
["wide"](https://en.wikipedia.org/wiki/Wide_and_narrow_data) format, 
whereas `drinks_smaller_tidy` is formatted in what's known as 
["long/narrow"](https://en.wikipedia.org/wiki/Wide_and_narrow_data#Narrow) format.

In the context of doing data science in R, 
long/narrow format \index{long data format} is also known as "tidy" format. 
In order to use the `ggplot2` and `dplyr` packages 
for data visualization and data wrangling, 
your input data frames *must* be in "tidy" format. 
Thus, all non-"tidy" data must be converted to "tidy" format first. 
Before we convert non-"tidy" data frames 
like `drinks_smaller` to "tidy" data frames like `drinks_smaller_tidy`, 
let's define "tidy" data.


### Definition of "tidy" data {#tidy-definition}

You have surely heard the word "tidy" in your life:

* "Tidy up your room!"
* "Write your homework in a tidy way so it is easier to provide feedback."
* Marie Kondo's best-selling book, 
  [_The Life-Changing Magic of Tidying Up: The Japanese Art of Decluttering and Organizing_](https://www.powells.com/book/-9781607747307), 
  and Netflix TV series [_Tidying Up with Marie Kondo_](https://www.netflix.com/title/80209379).
* "I am not by any stretch of the imagination a tidy person, 
  and the piles of unread books on the coffee table and by my bed 
  have a plaintive, pleading quality to me - 'Read me, please!'" - Linda Grant

What does it mean for your data to be "tidy"? 
While "tidy" has a clear English meaning of "organized," 
the word "tidy" in data science using R means 
that your data follows a standardized format. 
We will follow Hadley Wickham's \index{Wickham, Hadley} definition 
of *"tidy" data* \index{tidy data} [@tidy] 
shown also in Figure \@ref(fig:tidyfig):

(ref:tidy-r4ds) Tidy data graphic from *R for Data Science*.

```{r tidyfig, echo=FALSE, fig.cap="(ref:tidy-r4ds)", out.height="80%", out.width="80%", purl=FALSE}
knitr::include_graphics(here::here(
                                   "docs", 
                                   "images", 
                                   "r4ds", 
                                   "tidy-1.png")
)
```

> 
  A *dataset* is a collection of values, 
  usually either numbers (if quantitative) 
  or strings AKA text data (if qualitative/categorical). 
  Values are organised in two ways. 
  Every value belongs to a variable and an observation. 
  A variable contains all values that measure the same underlying attribute 
  (like height, temperature, duration) across units. 
  An observation contains all values measured on the same unit 
  (like a person, or a day, or a city) across attributes.
> 
  "Tidy" data is a standard way of mapping the meaning of a dataset 
  to its structure. 
  A dataset is messy or tidy depending on how rows, columns and tables 
  are matched up with observations, variables and types. 
  In *tidy data*:
>
> 1. Each variable forms a column.
> 2. Each observation forms a row.
> 3. Each type of observational unit forms a table.


For example, say you have the following table of stock prices 
in Table \@ref(tab:non-tidy-stocks):

```{r non-tidy-stocks, echo=FALSE, purl=FALSE}
stocks <- tibble::tibble(
  Date = as.Date("2009-01-01") + 0:4,
  `Boeing stock price` = paste("$", c("173.55", "172.61", "173.86", "170.77", "174.29"), sep = ""),
  `Amazon stock price` = paste("$", c("174.90", "171.42", "171.58", "173.89", "170.16"), sep = ""),
  `Google stock price` = paste("$", c("174.34", "170.04", "173.65", "174.87", "172.19"), sep = "")
) %>%
  dp$slice(1:2)
stocks %>%
  knitr::kable(
    digits = 2,
    caption = "Stock prices (non-tidy format)",
    booktabs = TRUE,
    linesep = ""
  ) %>%
  kableExtra::kable_styling(
    font_size = 16, 
    latex_options = c("hold_position")
  )
```

Although the data are neatly organized in a rectangular spreadsheet-type format, 
they do not follow the definition of data in "tidy" format. 
Although there are three variables 
corresponding to three unique pieces of information 
(date, stock name, and stock price), 
the three variables are not organized in three columns. 
In "tidy" data format, each variable should be its own column, 
as shown in Table \@ref(tab:tidy-stocks). 
Notice that both tables present the same information, 
but in different formats. 

```{r tidy-stocks, echo=FALSE, purl=FALSE}
stocks_tidy <- stocks %>%
  dp$rename(
    Boeing = `Boeing stock price`,
    Amazon = `Amazon stock price`,
    Google = `Google stock price`
  ) %>%
  #  gather(`Stock name`, `Stock price`, -Date)
  tidyr::pivot_longer(
    cols = -Date,
    names_to = "Stock Name",
    values_to = "Stock Price"
  )
stocks_tidy %>%
  knitr::kable(
    digits = 2,
    caption = "Stock prices (tidy format)",
    booktabs = TRUE,
    linesep = ""
  ) %>%
  kableExtra::kable_styling(
    font_size = 16, 
    latex_options = c("hold_position")
  )
```

Now we have the requisite three columns Date, Stock Name, and Stock Price. 
On the other hand, consider the data in Table \@ref(tab:tidy-stocks-2).

```{r tidy-stocks-2, echo=FALSE, purl=FALSE}
stocks <- tibble::tibble(
  Date = as.Date("2009-01-01") + 0:4,
  `Boeing Price` = paste("$", c("173.55", "172.61", "173.86", "170.77", "174.29"), sep = ""),
  `Weather` = c("Sunny", "Overcast", "Rain", "Rain", "Sunny")
) %>%
  dp$slice(1:2)
stocks %>%
  knitr::kable(
    digits = 2,
    caption = "Example of tidy data" # ,
    #    booktabs = TRUE
  ) %>%
  kableExtra::kable_styling(
    font_size = 16, 
    latex_options = c("hold_position")
  )
```

In this case, even though the variable "Boeing Price" 
occurs just like in our non-"tidy" data in Table \@ref(tab:non-tidy-stocks), 
the data *is* "tidy" because there are three variables 
corresponding to three unique pieces of information: Date, Boeing price, 
and the Weather on that particular day.

```{block, type="learncheck", purl=FALSE}
\vspace{-0.15in}
**_Learning check_**
\vspace{-0.1in}
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
What are common characteristics of "tidy" data frames?

<!-- **`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`**  -->
<!-- What makes "tidy" data frames useful for organizing data? -->

```{block, type="learncheck", purl=FALSE}
\vspace{-0.25in}
\vspace{-0.25in}
```


### Converting to "tidy" data

In this book so far, you've only seen data frames 
that were already in "tidy" format. 
Furthermore, for the rest of this book, 
you'll mostly only see data frames that are already in "tidy" format as well. 
This is not always the case however with all datasets in the world. 
If your original data frame is in wide (non-"tidy") format 
and you would like to use the `ggplot2` or `dplyr` packages, 
you will first have to convert it to "tidy" format. 
To do so, we recommend using the \index{tidyr!pivot\_longer()} `pivot_longer()` 
function in the `tidyr` \index{R packages!tidyr} package [@R-tidyr]. 

Going back to our `drinks_smaller` data frame from earlier:

```{r}
drinks_smaller
```

We can convert it to "tidy" format by using the `pivot_longer()` function 
from the `tidyr` package as follows:

```{r}
drinks_smaller_tidy <- drinks_smaller %>% 
  tidyr::pivot_longer(names_to = "type", 
               values_to = "servings", 
               cols = -country)
drinks_smaller_tidy
```

We set the arguments to `pivot_longer()` as follows:

1. `names_to` here corresponds to the name of the variable in the new "tidy"/long data frame that will contain the *column names* of the original data. Observe how we set `names_to = "type"`. In the resulting `drinks_smaller_tidy`, the column `type` contains the three types of alcohol `beer`, `spirit`, and `wine`. Since `type` is a variable name that doesn't appear in `drinks_smaller`, we use quotation marks around it. You'll receive an error if you just use `names_to = type` here.
1. `values_to` here is the name of the variable in the new "tidy" data frame that will contain the *values* of the original data. Observe how we set `values_to = "servings"` since each of the numeric values in each of the `beer`, `wine`, and `spirit` columns of the `drinks_smaller` data corresponds to a value of `servings`. In the resulting `drinks_smaller_tidy`, the column `servings` contains the `r n_countries` $\times$ `r n_alcohol_types` = `r n_row_drinks` numerical values. Note again that `servings` doesn't appear as a variable in `drinks_smaller` so it again needs quotation marks around it for the `values_to` argument.
1. The third argument `cols` is the columns in the `drinks_smaller` data frame 
   you either want to or don't want to "tidy." 
   Observe how we set this to `-country` indicating that 
   we don't want to "tidy" the `country` variable in `drinks_smaller`. 
   As a result, only `beer`, `spirit`, and `wine` columns from the original 
   data frame will be transformed. 
   Because `country` is an existing column 
   from the original data frame `drinks_smaller`, 
   we don't put quotation marks around it.

The argument `cols` can be specified in more than one way. 
Let's consider code that's written slightly differently 
but that produces the same output: 

```{r eval=FALSE}
drinks_smaller %>% 
  tidyr::pivot_longer(names_to = "type", 
               values_to = "servings", 
               cols = c(beer, spirit, wine))
```

Note that the third argument now specifies which columns we want to "tidy": 
`c(beer, spirit, wine)`, 
instead of the columns we don't want to "tidy": `-country`. 
We use the `c()` function to create a vector of the columns in `drinks_smaller` 
that we'd like to "tidy." 
Note that because these three columns are adjacent to each other, 
in the `drinks_smaller` data frame, 
we could also do the following for the `cols` argument:

```{r eval=FALSE}
drinks_smaller %>% 
  tidyr::pivot_longer(names_to = "type", 
               values_to = "servings", 
               cols = beer:wine)
```

With our `drinks_smaller_tidy` "tidy" formatted data frame, we can now produce the barplot you saw in Figure  \@ref(fig:drinks-smaller) using `geom_col()`. This is done in Figure \@ref(fig:drinks-smaller-tidy-barplot). Recall from Section \@ref(geombar) on barplots that we use `geom_col()` and not `geom_bar()`, since we would like to map the "pre-counted" `servings` variable to the `y`-aesthetic of the bars.

```{r eval=FALSE}
gg$ggplot(drinks_smaller_tidy, gg$aes(x = country, y = servings, fill = type)) +
  gg$geom_col(position = "dodge")
```

(ref:drinks-col) Comparing alcohol consumption in `r n_countries` countries using geom_col().

```{r drinks-smaller-tidy-barplot, echo=FALSE, fig.cap="(ref:drinks-col)", purl=FALSE}
drinks_smaller_tidy_plot
```

Converting "wide" format data to "tidy" format often confuses new R users. 
The only way to learn to get comfortable with the `pivot_longer()` function 
is with practice, practice, and more practice using different datasets. 
For example, run `?tidyr::pivot_longer` 
and look at the examples in the bottom of the help file. 
We'll show another example of using `pivot_longer()` 
to convert a "wide" formatted data frame to "tidy" format 
in Section \@ref(case-study-tidy). 

If however you want to convert a "tidy" data frame to "wide" format, 
you will need to use the `pivot_wider()`\index{tidyr!pivot\_wider()} function instead. 
Run `?tidyr::pivot_wider` and look at the examples 
in the bottom of the help file for examples.

You can also view examples of both `pivot_longer()` and `pivot_wider()` 
on the [tidyverse.org](https://tidyr.tidyverse.org/dev/articles/pivot.html#pew) webpage. 
There's a nice example to check out the different functions available for data tidying and a case study using data from the World Health Organization on that webpage. Furthermore, each week the R4DS Online Learning Community posts a dataset in the weekly [`#`TidyTuesday event](https://github.com/rfordatascience/tidytuesday) that might serve as a nice place for you to find other data to explore and transform. 

```{block, type="learncheck", purl=FALSE}
\vspace{-0.15in}
**_Learning check_**
\vspace{-0.1in}
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
Take a look at the `airline_safety` data frame 
included in the `fivethirtyeight` data package. 
Run the following:

```{r purl=FALSE}
import::from(fivethirtyeight, df_airline_safety = airline_safety)
df_airline_safety
```

After reading the help file by running `?fivethirtyeight::airline_safety`, 
we see that `airline_safety` is a data frame 
containing information on different airline companies' safety records. 
This data was originally reported on the data journalism website, 
FiveThirtyEight.com, in Nate Silver's article, 
["Should Travelers Avoid Flying Airlines That Have Had Crashes in the Past?"](https://fivethirtyeight.com/features/should-travelers-avoid-flying-airlines-that-have-had-crashes-in-the-past/). 
Let's only consider the variables `airlines` 
and those relating to fatalities for simplicity:

```{r}
airline_safety_smaller <- df_airline_safety %>% 
  dp$select(airline, dp$starts_with("fatalities"))
airline_safety_smaller
```

This data frame is not in "tidy" format. How would you convert this data frame to be in "tidy" format, in particular so that it has a variable `fatalities_years` indicating the incident year and a variable `num` of the fatality counts?

```{block, type="learncheck", purl=FALSE}
\vspace{-0.25in}
\vspace{-0.25in}
```





### Data frames in `nycflights13`

Recall the `nycflights13` package 
we introduced in Section \@ref(nycflights13) 
with data about all domestic flights departing from New York City in 2013. 

```{r eval=F}
import::from(nycflights13, df_flights = flights)
```


Let's revisit the `df_flights` data frame by running `View(df_flights)`. 
We saw that `df_flights` has a rectangular shape, 
with each of its `r df_flights %>% nrow() %>% formatC(big.mark=",")` 
rows corresponding to a flight and each of its `r df_flights %>% ncol()` 
columns corresponding to different characteristics/measurements of each flight. 
This satisfied the first two criteria of the definition 
of "tidy" data from Subsection \@ref(tidy-definition): 
that "Each variable forms a column" and "Each observation forms a row." 
But what about the third property of "tidy" data 
that "Each type of observational unit forms a table"?

Recall that we saw in Subsection \@ref(exploredataframes) 
that the observational unit for the `df_flights` data frame 
is an individual flight. 
In other words, the rows of the `df_flights` data frame refer to characteristics/measurements of individual flights. Also included in the `nycflights13` package are other data frames with their rows representing different observational units [@R-nycflights13]:

```{r echo=F}
import::from(nycflights13, df_planes = planes)
import::from(nycflights13, df_weather = weather)
import::from(nycflights13, df_airlines = airlines)
```

* `airlines`: translation between two letter IATA carrier codes 
  and airline company names (`r df_airlines %>% nrow()` in total). 
  The observational unit is an airline company.
* `planes`: aircraft information about each of 
  `r df_planes %>% nrow() %>% formatC(big.mark=",")` planes used, 
  i.e., the observational unit is an aircraft.
* `weather`: hourly meteorological data 
  (about `r df_weather %>% dp$count(origin) %>% .[["n"]] %>% mean() %>% round() %>% formatC(big.mark=",")` 
  observations) for each of the three NYC airports, i.e., the observational unit is an hourly measurement of weather at one of the three airports.
* `airports`: airport names and locations. The observational unit is an airport.

The organization of the information into these five data frames follows the third "tidy" data property: observations corresponding to the same observational unit should be saved in the same table, i.e., data frame. You could think of this property as the old English expression: "birds of a feather flock together." 





## Case study: Democracy in Guatemala {#case-study-tidy}

In this section, we'll show you another example of how to convert a data frame that isn't in "tidy" format ("wide" format) to a data frame that is in "tidy" format ("long/narrow" format). We'll do this using the `pivot_longer()` function from the `tidyr` package again. 

Furthermore, we'll make use of functions from the `ggplot2` and `dplyr` packages to produce a *time-series plot* showing how the democracy scores have changed over the 40 years from 1952 to 1992 for Guatemala. Recall that we saw time-series plots in Section \@ref(linegraphs) on creating linegraphs using `geom_line()`. 

Let's use the `dem_score` data frame we imported in Section \@ref(csv), but focus on only data corresponding to Guatemala.

```{r}
guat_dem <- dem_score %>% 
  dp$filter(country == "Guatemala")
guat_dem
```

Let's lay out the grammar of graphics we saw in Section \@ref(grammarofgraphics). 

First we know we need to set `data = guat_dem` and use a `geom_line()` layer, but what is the aesthetic mapping of variables? We'd like to see how the democracy score has changed over the years, so we need to map:

* `year` to the x-position aesthetic and
* `democracy_score` to the y-position aesthetic

Now we are stuck in a predicament, much like with our `drinks_smaller` example in Section \@ref(tidy-data-ex). We see that we have a variable named `country`, but its only value is `"Guatemala"`.  We have other variables denoted by different year values.  Unfortunately, the `guat_dem` data frame is not "tidy" and hence is not in the appropriate format to apply the grammar of graphics, and thus we cannot use the `ggplot2` package just yet.  

We need to take the values of the columns corresponding to years in `guat_dem` and convert them into a new "names" variable called `year`. Furthermore, we need to take the democracy score values in the inside of the data frame and turn them into a new "values" variable called `democracy_score`. Our resulting data frame will have three columns:  `country`, `year`, and `democracy_score`. Recall that the `pivot_longer()` function in the `tidyr` package does this for us:

```{r}
guat_dem_tidy <- guat_dem %>% 
  tidyr::pivot_longer(names_to = "year", 
               values_to = "democracy_score", 
               cols = -country,
               names_transform = list(year = as.integer)) 
guat_dem_tidy
```

We set the arguments to `pivot_longer()` as follows:

1. `names_to` is the name of the variable in the new "tidy" data frame that will contain the *column names* of the original data. Observe how we set `names_to = "year"`.  In the resulting `guat_dem_tidy`, the column `year` contains the years where Guatemala's democracy scores were measured.
1. `values_to` is the name of the variable in the new "tidy" data frame that will contain the *values* of the original data. Observe how we set `values_to = "democracy_score"`. In the resulting `guat_dem_tidy` the column `democracy_score` contains the 1 $\times$ 9 = 9 democracy scores as numeric values.
1. The third argument is the columns you either want to or don't want to "tidy." Observe how we set this to `cols = -country` indicating that we don't want to "tidy" the `country` variable in `guat_dem` and rather only variables `1952` through `1992`. 
1. The last argument of `names_transform` tells R what type of variable `year` should be set to. Without specifying that it is an `integer` as we've done here, `pivot_longer()` will set it to be a character value by default.

We can now create the time-series plot in Figure \@ref(fig:guat-dem-tidy) to visualize how democracy scores in Guatemala have changed from 1952 to 1992 using a \index{ggplot2!geom\_line()} `geom_line()`. Furthermore, we'll use the `labs()` function in the `ggplot2` package to add informative labels to all the `aes()`thetic attributes of our plot, in this case the `x` and `y` positions.

```{r guat-dem-tidy, fig.cap="Democracy scores in Guatemala 1952-1992.", fig.height=3}
gg$ggplot(guat_dem_tidy, gg$aes(x = year, y = democracy_score)) +
  gg$geom_line() +
  gg$labs(x = "Year", y = "Democracy Score")
```

Note that if we forgot to include the `names_transform` argument specifying that `year` was not of character format, we would have gotten an error here since `geom_line()` wouldn't have known how to sort the character values in `year` in the right order. 

```{block lc-tidying, type="learncheck", purl=FALSE}
\vspace{-0.15in}
**_Learning check_**
\vspace{-0.1in}
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`**  Convert the `dem_score` data frame into
a "tidy" data frame and assign the name of `dem_score_tidy` to the resulting long-formatted data frame.

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** 
Read in the life expectancy data stored at 
<https://moderndive.com/data/le_mess.csv> 
and convert it to a "tidy" data frame. 
Hint: use `readr::read_csv()` function we learned earlier 
to import the data. 

```{block, type="learncheck", purl=FALSE}
\vspace{-0.25in}
\vspace{-0.25in}
```





## `tidyverse` package {#tidyverse-package}

So far I have managed to avoid discussing the elephant in the room: 
`tidyverse`. 
However, you must have seen it in Datacamp and many other references. 
What is `tidyverse`? 
It is an amalgamation of several packages, 
each specilizing on one aspect of data manipulation. 
So far, we have seen four of them in this course: 
`ggplot2` for data visualization, 
`dplyr` for data wrangling, 
`readr` for importing spreadsheet data into R, 
and `tidyr` for converting data to "tidy" format. 

The [`tidyverse`](https://www.tidyverse.org/) "umbrella" package 
gets its name from the fact that all the functions in all its packages are designed to have common inputs and outputs: data frames are in "tidy" format. This standardization of input and output data frames makes transitions between different functions in the different packages as seamless as possible. 

`tidyverse` was created mainly for convience purposes. 
Instead of having to install many different packages, 
one can install `tidyverse` and have access to all of them. 
However, I have chosen to stick with individual packages for this course. 
This choice was partly inspired by my own experience: 
it is easier to remember what each function does 
once I have learned their grouping under their respective packages 
(remember "chunking" from your psyc101?). 

Sooner or later, you will encounter R scripts 
that start with `install.packages("tidyverse")`, 
or `library(tidyverse)`. 
I recommend that you avoid using this umbrella package in your own code 
and always look functions up whenever you are unsure which packages they belongs to. 
Do so until you can comfortably identify the membership 
of most commonly used functions 
and are used to resorting to `help` whenever you are unsure. 




## Conclusion {#tidy-data-conclusion}

### Additional resources


### What's to come?

Congratulations! You've completed the "Data Science with `tidyverse`" portion of this book. We'll now move to the "Data modeling with moderndive" portion of this book in Chapters \@ref(simple-regression) and \@ref(multiple-regression), where you'll leverage your data visualization and wrangling skills to model relationships between different variables in data frames. 

However, we're going to leave Chapter \@ref(inference-for-regression) on "Inference for Regression" until after we've covered statistical inference in Chapters \@ref(clt), \@ref(confidence-intervals), and \@ref(hypothesis-testing). Onwards and upwards into Data Modeling as shown in Figure \@ref(fig:part2)!

(ref:flowchart-partii) *ModernDive* flowchart - on to Part II!

```{r part2, echo=FALSE, fig.cap="(ref:flowchart-partii)", fig.height=5.5, purl=FALSE}
knitr::include_graphics(here::here(
                                   "docs", 
                                   "images", 
                                   "flowcharts", 
                                   "flowchart", 
                                   "flowchart.005.png")
)
```
