--- 
title: "Baby Modern Dive"
author: "Chunyun Ma, Chester Ismay, Albert Y. Kim"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [bib/teaching_2021W.bib, bib/book.bib, bib/packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: chunyunma/baby-modern-dive
description: 'Adapted Modern Dive for CGSC5101\/HCIN5400'
url: 'https\://.chunyunma.me'
---


```{r load-package, message=F, warning=F}
#
# Install packages
#

cran_packages = c(
                  # Packages required by students: 
                  # "knitr", "ggplot2", 
                  # Packages used internally: 
                  # "bookdown", 
                  # "import", 
                  # "rmarkdown", "here", "remotes", "kableExtra"
)

# Need to cite these packages separately
gh_repos = c(
             # Required by students:
             
             # Used internally: 
             # "rapporter/pander", "dcomtois/summarytools"
             )

# TODO: maybe separate these into for_reader and for_authors

if (length(cran_packages) != 0) xf$pkg_load2(cran_packages)
if (length(gh_repos) != 0) xf$install_github(gh_repos)

#
# namespace
#
gp <- import::from(ggplot2, .all=TRUE, .into={new.env()})
kn <- import::from(knitr, .all=TRUE, .into={new.env()})
```

```{r knitr, message=F, warning=F}
#
# knitr options
#
kn$opts_chunk$set(
  tidy = FALSE,
  out.width = "\\textwidth",
  fig.align = "center",
  comment = NA
)

#
# Set output options
#
if (kn$is_html_output()) {
  options(width = 80)
}

#
# automatically create a bib database for R packages
#
kn$write_bib(
                 c('xfun', 
                   cran_packages
                   ), 
                 here::here('bib', 'packages.bib')
)
```

```{r ggplot}
#
# Set ggplot2 theme
#
gp$theme_set(gp$theme_grey())
```


# Welcome {.unnumbered #welcome}

This is the website for the course notes of CGSC5101/HCIN5400. 

## What's with the name? {-}

The material is largely based on a book called "Modern Dive". 
<!-- may expand the list of inspirations -->
I first stumbled upon the book 
when I was preparing for teaching this course in 2019. 
I personally benefited greatly from reading the book, 
and teaching a course using its material.

## Who are the players? {-}

### You professor {-}

My background

My history with teaching this course and teaching stats in general

#### Why choosing R {-}

#### What do I hope to deliver, content-wise {-}

<!-- stats vs. data science 
see "What we hope you will learn from this hope in Preface"
A modified version:
Experience the life cycle of a data analysis pipeline 
from raw data to meaninful narration, ready for readers 
of various background. This entails: 
- data wrangling (maybe a different term is better)
- Apply appropriate model to the data
- "decode" output filled with technical jargons, interpret results, make inferences
- Learn how to use R to turn the process reproducible
There are more minor objectives, but these should give a high-level overview

Elaborate on reproducible research
-->

### Your teaching assistant {-}

### Your peers {-}

## Playground rules {-}

### Communication policy {-}

What worked last term

### Minimize distraction {-}

### Structure of each lecture {-}

theory + hands-on


### Syllabus & assignments {-}


### Datacamp {-}

