---
title: "Data challenge"
author: "First Last"
date: "2021-02-12"
output: 
    html_document:
        number_sections: true
---


# Problem statement [^1]

At Shopify, there are 100 sneaker shops on line, 
and each of these shops sells only one model of shoe. 
We want to do some analysis of the average order value (AOV) [^2] 
across all the 100 sneaker shops. 
When we look at orders data over a 30 day window, 
we naively calculate an AOV of $3145.13. 
Given that we know these shops are selling sneakers, 
a relatively affordable item, something seems wrong with our analysis.
Use the following three questions to guide your investigation. 

(a) Investigate what could have gone wrong with this data. 

(b) What alternate statistics would you report for this dataset?

(c) What is its value (their values)?


# Import the data


```{r packages, message = FALSE, echo=F}
# Install xfun so that I can use xfun::pkg_load2
if (!requireNamespace('xfun')) install.packages('xfun')
xf <- loadNamespace('xfun')

cran_packages = c(
                  "dplyr", 
                  "ggplot2", 
                  "knitr", 
                  "readr", 
                  "tibble"
                  )
if (length(cran_packages) != 0) xf$pkg_load2(cran_packages)

import::from(magrittr, '%>%')
gg <- import::from(ggplot2, .all=TRUE, .into={new.env()})
dp <- import::from(dplyr, .all=TRUE, .into={new.env()})
```


```{r message=F}
df_sales <- readr::read_csv("https://query.data.world/s/e76we3vpgut5cmydqgm2nb7e5xqx7i", 
                            col_types = readr::cols(
                                             order_id = readr::col_integer(), 
                                             shop_id = readr::col_integer(), 
                                             user_id = readr::col_integer(), 
                                             order_amount = readr::col_double(), 
                                             total_items = readr::col_integer(), 
                                             payment_method = readr::col_factor(), 
                                             created_at = readr::col_datetime(format = "%Y-%m-%d %H:%M:%z")
                                             )
)
# df_sales # sanity check to make sure the data has been imported correctly
```

Most of the variables in the dataset are self-explanatory. 
Nevertheless, here is a list of brief descriptions for each one. 

+ `order_id`: A unique identifier for each order placed by a customer online
+ `shop_id`: A unique identifier for the sneaker shop involved in 
  an online order
+ `order_amount`: The amount of money each online order incurred
+ `total_items`: The number of sneakers purchased in each online order
+ `payment_method`: The type of payment method used in an online order, 
  e.g., credit card, cash.
+ `created_at`: The time stamp when an online order was created


# Eyeball the data

```{r}
tibble::glimpse(df_sales)
```

From the print out, can you tell how many rows and columns are in the data? 
(5000 rows/transactions, and 7 columns)

Print out the first three rows to get an idea of what the data actually look like. 

```{r}
head(df_sales, 3) %>% 
  knitr::kable(
               caption = "First three rows of the `df_sales` data frame"
  )
```

Confirm where the value $3145.13 in the problem statement came from. 

```{r}
mean(df_sales$order_amount)
```

<!-- Complete the report with your response.
Keep in mind that there are more than one way to solve the problem.
The following two headings are just an example 
of how to organize your final deliverable.
You may change them to your liking. -->

# Check for extreme values

Now we examine the minimum and maximum of `order_amount` and `total_items.` 
The dubious result is likely due to some extreme values 
in either or both of the two varables.


<!-- Your response below -->


# Summary

<!-- Your response below -->


[^1]: This problem was based on the Shopify Data Science Challenge in 2019. 

[^2]: Average Order Value (AOV) means average dollar spent per order. 
