---
title: ""  # swap with your own
author: "Your Name"
date: "year-month-date"
output: html_document
---

# Preamble (delete this section before submitting your work)

- If there are a large number variables in your dataset, 
  some of the analyses listed below may become tedious. 
  Try to group them when describing the fingings, 
  especially if some variables are of the same type (e.g., 5 point likert scale).

- This guide tries to provide a one-size-fits-all solution to descriptives, 
  which is an impossible task to begin with. 
  I have merely listed some typical analyses 
  you may encounter during this stage of data analysis. 
  *Not every prompt included in this template would apply to your case.*
  *Choose ones that you think are relevant.*
  Other than answering the questions mentioned in this template, 
  you may want to take a look at the original paper, 
  and consider what descriptives have been reported there. 
  Not sure what counts as descriptives (vs. inferential statistics)? 
  Look for these keywords: 
  number of participants, 
  their mean/median age, 
  male vs. female ratio, 
  percentages, 
  demographics, etc.

- Please feel free to re-structure this document as you see fit. Own it!

- Submit this Rmarkdown file as your deliverable

# Descriptives

```{r setup, include=FALSE}
knitr::opts_chunk$set(
                      # set any chunk options here
)
```

```{r packages, echo=FALSE}
cran_internal <- c(
                   # adjust this list based on your needs
                   "dplyr", 
                   "ggplot2", 
                   "skimr",
                   "tibble"
)
gg <- import::from(ggplot2, .all=TRUE, .into={new.env()})
dp <- import::from(dplyr, .all=TRUE, .into={new.env()})

import::from(magrittr, '%>%')
```

```{r import-data}
# import the dataset you will be working with and save it as a data frame
# this should be the dataset post cleaning/filtering

```



## Eye-ball the data

```{r}
# replace xx with the actual name of your dataframe
tibble::glimpse(xx)
```


<!-- - Include a description of each variable, in plain English.  -->
<!--   For example,  -->
<!--   *lifeExp – Average life expectancy in each country, measured in years.*  -->
<!--   In the description, you may address: -->
<!--  -->
<!--     - What does this variable measure? -->
<!--  -->
<!--     - Is this a numerical/continous variable or a categorical varialbe/factor? -->
<!--  -->
<!--     - If this is a numerical variable, what is the unit of the measurement? -->
<!--  -->
<!--     - Is this a variable you plan to use it in the replication analyses?  -->
<!--       This question is intended for those of you  -->
<!--       who may be working with a dataset which has more variables than what you need. -->
<!--       But if you still have many variables in this data drame  -->
<!--       that are irrelevant to your analysis,  -->
<!--       then you might want to go back to the data cleaning stage. -->



## Missing data analysis

```{r}
# Insert code below



```

<!-- For each key variable,  -->
<!-- investigate whether there is any missing value  -->
<!-- (empty cell, or NA, or any other types of illegitimate values).  -->
<!-- Report missingness on each variable: how many values are missing in each column. -->


## Numerical variables

<!-- For each numerical variable: -->
<!--  -->
<!-- - Calculate mean, standard deviation, minimum, maximum -->
<!--  -->
<!-- - Plot histogram and boxplot -->

```{r}
# Insert code below

```


<!-- Discuss any notable findings. 
For example, is the distribution of any variable clearly not normal? -->




## Categorical variables

<!-- For each categorical variable/factor: -->
<!--  -->
<!-- - List levels of this variable -->
<!--  -->
<!-- - Provide a frequency table -->

```{r}
# Insert code below

```



<!-- Discuss any notable findings. 
For example, is the frequency of observations 
for a certain level on a variable particularly low, e.g., smaller than 5? -->


