---
title: Cheat Sheet
output: 
  html_document: 
    toc: true
    toc_float: 
        collapsed: false
        smooth_scroll: false
---

```{r setup, include=F}
knitr::opts_chunk$set(
                      tidy=F, 
                      eval=F
)
```

```{r packages-and-data, include=F}
gg <- import::from(ggplot2, .all=T, .into={new.env()})
import::from(nycflights13, df_flights = flights)
import::from(magrittr, "%>%")
```


## Visualization

### Scatterplots

AKA bivariate plots. 
They allow you to visualize the *relationship* between two **continuous** variables. 

```{r}
# import the data frame and name it “flights”
import::from(nycflights13, df_flights = flights)

# choose all rows related to Alaska Airlaine carrier
df_alaska_flights <- df_flights %>% 
  dplyr::filter(carrier == "AS")

# built a scatter plot
gg$ggplot(data = df_alaska_flights, 
          mapping = gg$aes(x = dep_delay, y = arr_delay)) +
  gg$geom_point()
```

### Overplotting

When points are being plotted on top of each other over and over again 
and it is difficult to know the number of points being plotted.


+ Method 1: Changing the transparency of the points

Setting the `alpha` argument in `geom_point` 
(usually, `alpha` argument is set by default at 1 point – 100% opaque). 
By specifying a value of `alpha` argument, 
we can change the transparency of the points (to less than 1).

```{r demo-alpha eval=T}
df_alaska_flights <- df_flights %>% 
  dplyr::filter(carrier == "AS")

gg$ggplot(data = df_alaska_flights, 
          mapping = gg$aes(x = dep_delay, y = arr_delay)) +
  gg$geom_point(alpha = 0.2)
```

